### 2016-01-26 presentation/presentation_spanner.tex
Initial commit

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Beamer Presentation
% LaTeX Template
% Version 1.0 (10/11/12)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%-------------------------------------------------------------------------------
%	PACKAGES AND THEMES
%-------------------------------------------------------------------------------

\documentclass{beamer}

\begin{filecontents}{\jobname.bib}

\end{filecontents}

\mode<presentation> {

% The Beamer class comes with a number of default slide themes
% which change the colors and layouts of slides. Below this is a list
% of all the themes, uncomment each in turn to see what they look like.
\usetheme{Madrid}
}

\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{color, colortbl}
\usepackage{natbib}
\usepackage{setspace}
\usepackage[labelformat=simple]{caption}
\usepackage{tikz}

\usetikzlibrary{arrows,positioning,calc}

\definecolor{Lightgray}{rgb}{0.9, 0.9, 0.9}

%-------------------------------------------------------------------------------
%	TITLE PAGE
%-------------------------------------------------------------------------------

\title[]{Spanner: Google's Globally-Distributed Database}
\subtitle[]{}

\author{Kocher Daniel}

\institute[PLUS] 
{
Supervisor: Dipl.-Ing. Nikolaus Augsten, PhD \newline \newline \newline
University of Salzburg \\ 
\medskip
Daniel.Kocher@stud.sbg.ac.at
}
\date{\today}

\begin{document}
%-------------------------------------------------------------------------------
\begin{frame}
\titlepage
\end{frame}

%-------------------------------------------------------------------------------
\begin{frame}
\frametitle{Outline} 
\tableofcontents[hideallsubsections]
\end{frame}

%-------------------------------------------------------------------------------
\section{Motivation}
\begin{frame}
	\frametitle{Motivation}
	\begin{itemize}
    \item{Distributed data at global scale}
    \item{Externally-consistent distributed transactions}
    \item{High availability}
    \item{Two "competitors": \emph{Bigtable} \& \emph{Megastore}}
    \pause
    \item{Why not use Bigtable?}
    \begin{itemize}
      \item{Difficult to use for applications with complex, evolving schemas}
      \item{Only \emph{eventually-consistent} (no strong consistency)}
    \end{itemize}
    \pause
    \item{Why not use Megastore}
    \begin{itemize}
      \item{Poor write throughput}
    \end{itemize}

  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\begin{frame}
  \frametitle{Characteristics}
  \begin{center}
    \textbf{Spanner} evolved from a Bigtable-like versioned key-value store to a
    \textbf{temporal multi-version database}
  \end{center}
  \pause
  \begin{itemize}
    \item{Versioned data is stored in schematized semi-relational tables}
    \item{Each version is automatically timestamped with its commit time}
    \item{Garbage Collection \& Read at old timestamps}
    \item{\textbf{General-purpose transactions}}
    \item{SQL-like query language}
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\begin{frame}
  \frametitle{Fundamentals}
  \begin{itemize}
    \item{Transaction $i$ is denoted $T_i$}
    %\item{RW \ldots Read-Write (transaction)}
    %\item{RO \ldots Read-Only (transaction)}
    \item{\textbf{Externally-consistent:} \newline
      If $T_1$ preceeds $T_2$ (Commit of $T_1$ happens before Begin of $T_2$, no
      overlap), then $T_1$ is serialized first
    }
    \pause
    \item{\textbf{Two-Phase Locking (2PL):} \newline
    Guarantees serializability
    \begin{enumerate}
      \item{\emph{Expanding} phase: locks are aquired, no locks are released}
      \item{\emph{Shrinking} phase: locks are released, no locks are aquired}
    \end{enumerate}
    }
    \item{\textbf{Two-Phase Commit (2PC):} \newline
    Coordinates processes participating in atomic distributed transaction
    \begin{enumerate}
      \item{\emph{Commit-Request} phase: Request "Yes" (Commit) or "No" (Abort) from
        every transaction process
      }
      \item{\emph{Commit} phase: Commit transaction if all voted "Yes", otherwise abort}
    \end{enumerate}
    }
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\begin{frame}
  \frametitle{Features}
  As a globally-distributed database, Spanner provides interesting features:
  \begin{itemize}
    \item{Dynamically controlled replication configurations at fine grain}
    \item{Externally consistent reads \& writes}
    \item{globally-consistent reads at a timestamp}
  \end{itemize}
  $\Rightarrow$ enables \textbf{atomic schema changes} in presence of ongoing
  transactions \newline
  \pause
  \begin{itemize}
    \item{Timestamps reflect serialization order (linearizability)}
    \item{Key enabler: novel TrueTime API exposing clock uncertainty}
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\section{Implementation}
\begin{frame}
  \frametitle{Implementation (Top-down)}
  \only<1-3>{
    \begin{itemize}
      \item{Spanner deployment is called \textbf{universe}}
      \item{Organized as a set of \textbf{zones}}
      \item{Zones are the unit of administrative deployment \& physical isolation}
      \item{Set of zones is the set of locations for replication}
      \pause
      \item{Each zone consists of}
      \begin{itemize}
        \item{several \textbf{spanservers} to store the data}
        \item{a \textbf{zonemaster} to assign data to spanservers}
        \item{a \textbf{location proxy} for clients to locate the assigned
          spanservers
        }
      \end{itemize}
      \pause
      \item{\textbf{universe master:} console to display status information about zones}
      \item{\textbf{placement driver:}}
      \begin{itemize}
        \item{handles automated data movement across zones}
        \item{moves data due to updated replication constraints or load balancing}
      \end{itemize}
    \end{itemize}
  }
  \only<4>{
    \begin{figure}
      \centering
      \includegraphics[width=.8\textwidth]{figures/spanner-server-organization.png}
      \caption{Spanner server organization}
    \end{figure}
  }
\end{frame}

%-------------------------------------------------------------------------------
\subsection{Spanserver Software Stack}
\begin{frame}
  \frametitle{Spanserver Software Stack}
  \only<1>{
    \begin{itemize}
      \item{Responsible for 100 - 1000 \textbf{tablets}}
      \item{A tablet is a bag of key-value mappings \newline
        \texttt{(key:string, timestamp:int64) $\rightarrow$ string}
      }
      \item{Timestamp are assigned to data $\Rightarrow$
        \textbf{multi-version database}
      }
      \item{Tablet states are stored on Colossus (successor of the GFS)}
      \item{Single Paxos state machine on top of each tablet to $\Rightarrow$
        \textbf{Replication}
      }
      \item{Set of replicas is called \textbf{Paxos group}}
    \end{itemize}
  }
  \pause
  \only<2>{
    \begin{itemize}
      \item{Every replica which is a leader implements}
      \begin{itemize}
        \item{a \textbf{lock table} for concurrency control: \texttt{(key range) $\rightarrow$ lock state}}
        \item{a \textbf{transaction manager} (TM) for distributed transactions}
      \end{itemize}
      \item{Transaction manager used to implement \textbf{participant leader}}
      \item{If multiple Paxos groups are involved: Two-Phase Commit}
      \begin{itemize}
        \item{One participant group is chosen as \textbf{coordinator}}
        \item{Participant leader of this group: \textbf{coordinator leader}}
      \end{itemize}
      \item{State of each Transaction manager is stored in Paxos group}
    \end{itemize}
  }
  \pause
  \only<3>{
    \begin{figure}
      \centering
      \includegraphics[width=.7\textwidth]{figures/spanserver-software-stack.png}
      \caption{Spanserver software stack}
    \end{figure}
  }
\end{frame}

%-------------------------------------------------------------------------------
\subsection{Directories \& Placement}
\begin{frame}
  \frametitle{Directories \& Placement}
  \only<1-4>{
    \begin{itemize}
      \item{\textbf{Directories:}}
      \begin{itemize}
        \item{Bucketing abstraction on top of key-value mappings}
        \item{Contain set of contiguous keys sharing common prefix}
        \item{Prefix allows to control locality of data}
        \item{Unit of data placement (same replication configuration)}
        \item{Smallest unit to specify geographic-replication properties (\textbf{placement})}
      \end{itemize}
      \pause
      \item{Data is moved directory-wise between Paxos groups}
      \item{Tablets may have multiple \textbf{partitions} of row space \newline
        $\Rightarrow$ co-locate directories frequently accessed together
      }
      \pause
      \item{Placements can be controlled in two dimensions:}
      \begin{enumerate}
        \item{Number \& types of replicas}
        \item{Geographic placement of replicas}
      \end{enumerate}
      \pause
      \item{\textbf{Movedir:} background task to move directories}
    \end{itemize}
  }
  \only<5>{
    \begin{figure}
      \centering
      \includegraphics[width=.6\textwidth]{figures/directories.png}
      \caption{Directories (the unit of data movement)}
    \end{figure}
  }
\end{frame}

%-------------------------------------------------------------------------------
\subsection{Data Model}
\begin{frame}
  \frametitle{Data Model}
  \only<1>{
    \begin{itemize}
      \item{Features:}
      \begin{itemize}
        \item{Schematized semi-relational tables}
        \item{SQL-like query language}
        \item{General-purpose transactions}
      \end{itemize}
      \item{Layered on top of directory-bucketed key-value mappings}
      \item{Applications create \textbf{databases} in universe}
      \item{Database may contain unlimited number of schematized \textbf{tables}}
      \item{Table contains rows (must have names), columns \& versioned values}
    \end{itemize}
  }
  \only<2>{
    \begin{figure}
      \centering
      \includegraphics[width=.5\textwidth]{figures/interleaving-example.png}
      \caption{Interleaving example}
    \end{figure}
    \begin{itemize}
      \item{Interleaving of tables to form directories is significant}
      \item{Locality relationships between multiple tables $\Rightarrow$ good
        performance
      }
    \end{itemize}
  }
\end{frame}

%-------------------------------------------------------------------------------
\section{TrueTime}
\begin{frame}
  \frametitle{TrueTime}
  \only<1>{
    \begin{table}
      \centering
      \begin{tabular}{|l||l|}
        \hline
        Method & Returns \tabularnewline
        \hline\hline
        $TT.now()$ & $TTinterval$: [$earliest$, $latest$] \tabularnewline
        \hline
        $TT.after(t)$ & true if $t$ has definitely passed \tabularnewline
        \hline
        $TT.before(t)$ & true if $t$ has definitely not arrived \tabularnewline
        \hline
      \end{tabular}
      \caption{TrueTime API ($t$ is of type $TTstamp$)}
    \end{table}
    Time is represented as a $TTinterval$ (interval with bounded uncertainty)
    \newline\newline
    Let $e$ be an event and $t_{abs}(e)$ denote the absolute time of $e$.
    \newline
    For $tt = TT.now()$, the following property holds:
    \begin{center}
      $tt.earliest \leq t_{abs}(e) \leq tt.latest$
    \end{center}
  }
  \only<2>{
    \begin{figure}
      \centering
      \begin{tikzpicture}[thick]
        \node (left-end) at (0, 0) {};
        \node[right = 6 of left-end] (right-end) {};

        \node[right = 1 of left-end] (left-bound) {};
        \node[left = 1 of right-end] (right-bound) {};

        \node[above = 0.5 of left-bound] (left-bound-top) {};
        \node[below = 0.5 of left-bound] (left-bound-bottom) {};
        \node[above = 0.5 of right-bound] (right-bound-top) {};
        \node[below = 0.5 of right-bound] (right-bound-bottom) {};

        \draw[blue!35, fill=blue!35] (left-bound-top.center) rectangle (right-bound-bottom.center);
        \node (two-e-invis) at ($(left-bound.south)!0.5!(right-bound.south)$) {};
        \node[below = 0 of two-e-invis.center] (two-e-t) {$2\epsilon$};
        \draw[->] (left-end.center) -- (right-end.center) node[right] {t};

        \draw (left-bound.center) -- (left-bound-top.center) -- ++(0.25, 0);
        \draw (left-bound.center) -- (left-bound-bottom.center) -- ++(0.25, 0);
        \draw (right-bound.center) -- (right-bound-top.center) -- ++(-0.25, 0);
        \draw (right-bound.center) -- (right-bound-bottom.center) -- ++(-0.25, 0);

        \node[above = 0 of left-bound-top.north] (left-bound-t) {earliest};
        \node[above = 0 of right-bound-top.north] (right-bound-t) {latest};

        \node[below = 0.75 of left-bound] (left-bound-e) {};
        \node[below = 0.75 of right-bound] (right-bound-e) {};

        \draw[<->] (left-bound-e.center) -- (right-bound-e.center);
        %\node[below right= 1 and 2.5 of left] {$2\epsilon$};
      \end{tikzpicture}
      \caption{Visualization of $TT.now()$ and $\epsilon$}
    \end{figure}
  }
  \only<3>{
    \begin{itemize}
      \item{Time references: GPS and atomic clocks (different failure modes)}
      \item{Set of \textbf{time master} machines per datacenter}
      \item{\textbf{Armageddon masters}: masters using atomic clocks}
      \item{\textbf{timeslave daemon} per machine}
      \pause
      \item{Masters compare their time references regularly}
      \item{Daemons poll a variety of masters (liar detection)}
      \pause
      \item{$\epsilon$ is the instantaneous error bound (typically 1 - 7ms)}
      \item{$\overline{\epsilon}$ is the average error bound (typically 4ms)}
      \item{$\epsilon$ is derived from applied worst-case local clock drift ($200\mu s/sec$)}
    \end{itemize}
  }
\end{frame}

%-------------------------------------------------------------------------------
\section{Concurrency Control}
\begin{frame}
  \frametitle{Concurrency Control}
  \begin{itemize}
    \item{TrueTime is used to guarantee correctness properties}
    \item{Properties are used to implement}
    \begin{itemize}
      \item{Externally-consistent transactions}
      \item{Lock-free read-only transactions}
      \item{Non-blocking reads in the past}
    \end{itemize}
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\subsection{Timestamp Management}
\begin{frame}
  \frametitle{Timestamp Management}
  \only<1>{
    \footnotesize
    \begin{table}
      \begin{tabular}{|l||l|l|}
        \hline
        Operation & CC & Replica required\tabularnewline
        \hline\hline
        Read-Write transaction & Pessimistic & leader\tabularnewline
        \hline
        Read-Only transaction & Lock-free & leader; any
        \tabularnewline
        \hline
        Snapshot Read, client-provided timestamp & Lock-free & any\tabularnewline
        \hline
        Snapshot Read, client-provided bound & Lock-free & any \tabularnewline
        \hline
      \end{tabular}
      \caption{Supported types of reads \& writes}
    \end{table}
  }
  \begin{itemize}
    \item{Standalone write $\Rightarrow$ RW transaction}
    \item{Non-snapshot standalone read $\Rightarrow$ RO transaction}
    \item{RO transactions}
    \begin{itemize}
      \item{must predeclare to not include writes}
      \item{can proceed on any sufficiently up-to-date-replica (also snapshot reads)}
    \end{itemize}
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\subsubsection{Paxos Leader Leases}
\begin{frame}
  \frametitle{Paxos Leader Leases}
  \begin{itemize}
    \item{Timed leases for long-lived leadership (10s by default)}
    \item{Quorum of lease votes $\Rightarrow$ leader has a lease}
    \item{Successful writes extend lease vote on replica}
    \item{\textbf{Disjointness invariant:} \newline
      For each Paxos group, each Paxos leader's \emph{lease interval} is disjoint
      from every other leader's
    }
    \item{Leaders must not abdicate before $TT.after(s_{max})$, where $s_{max}$
      denotes the maximum timestamp used by a leader
    }
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\subsubsection{Timestamps in RW Transactions}
\begin{frame}
  \frametitle{Timestamps in RW Transactions}
  \begin{itemize}
    \item{Timestamp assignment only in between the two phases (2PL)}
    \item{\textbf{Monotonicity variant:} \newline
      Within each Paxos group, Spanner assigns timestamps to Paxos writes in
      monotonically increasing order, even across leaders
    }
    \item{Enforced across leaders by disjointness invariant: \newline
      Leader must only assign timestamps within the interval of leader lease
    }
    \item{Timestamp $s$ assigned $\Rightarrow$ $s_{max} = s$}
    \item{\textbf{External consistency invariant}: \newline
      \begin{center}
        $t_{abs}\left(e^{commit}_1\right) < t_{abs}\left(e^{start}_2\right)$
        $\Rightarrow s_1 < s_2$
      \end{center}
    }
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\subsubsection{Serving Reads at a Timestamp}
\begin{frame}
  \frametitle{Serving Reads at a Timestamp}
  \begin{itemize}
    \item{Monotonicity invariant allows to determine if a replica's state is
      \textbf{sufficiently up-to-date} to satisfy a read
    }
    \item{\textbf{Safe time} $t_{safe}$ at each replica}
    \item{Replica satisfies read if $t \leq t_{safe}$}
    \item{$t_{safe} = min\left(t^{Paxos}_{safe}, t^{TM}_{safe}\right)$}
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\subsubsection{Timestamps in RO Transactions}
\begin{frame}
  \frametitle{Timestamps in RO Transactions}
  \begin{itemize}
    \item{Two phases:}
    \begin{enumerate}
      \item{Assign timestamp $s_{read}$}
      \item{Execute transaction's reads as snapshot reads at $s_{read}$}
    \end{enumerate}
    \item{Spanner assigns the oldest timestamp that preserves external consistency
      to reduce the changes of blocking
    }
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\subsection{Details}

\subsubsection{Read-Write Transactions}
\begin{frame}
  \frametitle{Read-Write Transactions}
  \begin{itemize}
    \item{RW transactions buffered at client until Commit}
    \item{Reads do not see effects of transaction's writes}
    \item{Client has completed all reads \& buffered all writes $\Rightarrow$ 2PC}
    \pause
    \item{Non-coordinator-participant leader}
    \begin{enumerate}
      %\item{Acquires write locks}
      \item{Chooses prepare timestamp according to monotonicity}
      \item{Logs prepare record through Paxos}
      \item{Notifies coordinator of its prepare timestamp}
    \end{enumerate}
    \pause
    \item{Coordinator leader}
    \begin{enumerate}
      %\item{Aquires write locks}
      \item{Chooses timestamp for whole transaction}
      \item{Commit timestamp $s$ must be}
      \begin{itemize}
        \item{$\geq$ all prepare timestamps}
        \item{$> TT.now().latest$ at the time the Commits were received}
        \item{$>$ any previously assigned timestamp}
      \end{itemize}
      \item{Logs Commit record through Paxos}
    \end{enumerate}
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\subsubsection{Read-Only Transactions}
\begin{frame}
  \frametitle{Read-Only Transactions}
  \begin{itemize}
    \item{\textbf{scope} expression summarizing the keys that will be read}
    \item{Served by single Paxos group \newline
      $\Rightarrow$ client issues transaction to group leader
    }
    \item{Served by multiple Paxos groups \newline
      $\Rightarrow$ client executes reads at $s_{read} = TT.now().latest$
    }
    \item{All reads can be sent to replicas that are sufficiently up-to-date}
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\subsubsection{Schema-Change Transactions}
\begin{frame}
  \frametitle{Schema-Change Transactions}
  \begin{itemize}
    \item{TrueTime enables this feature}
    \item{Use of standard transaction infeasible}
    \item{Non-blocking variant of standard transaction}
    \begin{enumerate}
      \item{Explicit assignment of a future timestamp (Prepare phase)}
      \item{Reads \& writes synchronize with any registered schema-change
        timestamp at time $t$
      }
    \end{enumerate}
    \item{Without TrueTime $\Rightarrow$ schema change at $t$ would be meaningless
    }
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
%\subsubsection{Refinements}
%\begin{frame}
%  \frametitle{Refinements}
%  \begin{itemize}
%    \item{}
%  \end{itemize}
%\end{frame}

%-------------------------------------------------------------------------------
\section{Evaluation}

\subsection{Microbenchmarks}
\begin{frame}
  \frametitle{Microbenchmarks}
  \only<1>{
    \begin{itemize}
      \item{Setup:}
      \begin{itemize}
        \item{4GB RAM scheduling units per spanserver}
        \item{4 cores (AMD Barcelona 2200MHz) per spanserver}
        \item{One spanserver per zone}
        \item{Clients \& zones in set of datacenters with network distance $<$ 1ms}
        \item{Database created with 50 Paxos groups with 2500 directories}
      \end{itemize}
      \item{Operations: standalone reads \& writes of 4KB}
      \item{One unmeasured round of reads to warm any location caches}
    \end{itemize}
  }
  \only<2>{
    \textbf{Latency [ms] (less is better)}
    \begin{table}
      \centering
      \begin{tabular}{|l||l|l|l|}
        \hline
        Replicas & Write & RO Transaction & Snapshot Read \tabularnewline
        \hline\hline
        1D & $9.4\pm0.6$ & - & - \tabularnewline
        \hline
        1 & $14.4\pm1.0$ & $1.4\pm0.1$ & $1.3\pm0.1$ \tabularnewline
        \hline
        3 & $13.9\pm0.6$ & $1.3\pm0.1$ & $1.2\pm0.1$ \tabularnewline
        \hline
        5 & $14.4\pm0.4$ & $1.4\pm0.05$ & $1.3\pm0.04$ \tabularnewline
        \hline
      \end{tabular}
      \caption{Latency experiments}
    \end{table}
    \begin{itemize}
      \item{Sufficiently few operations to avoid queuing}
      \item{Increasing number of replicas:}
      \begin{itemize}
        \item{Latency stays roughly constant with less standard deviation}
        \item{Latency to achieve quorum less sensitive to stragglers}
      \end{itemize}
    \end{itemize}
  }
  \only<3>{
    \textbf{Throughput [Kops/sec] (more is better)}
    \begin{table}
      \centering
      \begin{tabular}{|l||l|l|l|}
        \hline
        Replicas & Write & RO Transaction & Snapshot Read \tabularnewline
        \hline\hline
        1D & $4.0\pm0.3$ & - & - \tabularnewline
        \hline
        1 & $4.1\pm0.05$ & $10.9\pm0.4$ & $13.5\pm0.1$ \tabularnewline
        \hline
        3 & $2.2\pm0.5$ & $13.8\pm3.2$ & $38.5\pm0.3$ \tabularnewline
        \hline
        5 & $2.8\pm0.3$ & $25.3\pm5.2$ & $50.0\pm1.1$ \tabularnewline
        \hline
      \end{tabular}
      \caption{Throughput experiments}
    \end{table}
    \begin{itemize}
      \item{Sufficiently many operations to saturate servers' CPUs}
      \item{Increasing number of replicas:}
      \begin{itemize}
        \item{Snapshot read throughput increases almost linear}
      \end{itemize}
    \end{itemize}
  }
\end{frame}

%-------------------------------------------------------------------------------
\subsection{Availability}
\begin{frame}
  \frametitle{Availability}
  \only<1>{
    \begin{itemize}
      \item{Setup:}
      \begin{itemize}
        \item{Universe with 5 zones $Z_i$, each of which has 25 spanservers}
        \item{Database sharded into 1250 Paxos groups}
        \item{100 clients constantly issued snapshot reads (50K reads/sec)}
        \item{All leaders explicitly place in $Z_1$}
        \item{5s into each test, all servers in one zone were killed}
      \end{itemize}
    \end{itemize}
  }
  \only<2>{
    \begin{figure}
      \centering
      \includegraphics[width=.5\textwidth]{figures/effect-of-killing-servers-on-throughput.png}
      \caption{Effect of killing servers on throughput}
    \end{figure}
    \begin{itemize}
      \item{Availability benefits of running Spanner in multiple datacenters}
      %\item{Killing $Z_2$ (non-leader): no effect on read throughput}
      %\item{Killing $Z_1$ with notifications (leader-soft): minor effect}
      %\item{Killing $Z_1$ without notifications (leafer-hard): severe effect}
    \end{itemize}
  }
\end{frame}

%-------------------------------------------------------------------------------
\subsection{TrueTime}
\begin{frame}
  \frametitle{TrueTime}
  \begin{itemize}
    \item{Two questions:}
    \begin{itemize}
      \item{Is $\epsilon$ truly a bound on clock uncertainty?}
      \item{How bad does $\epsilon$ get?}
    \end{itemize}
    \pause
    \item{Answers:}
    \begin{itemize}
      \item{Clock issues infrequent relative to more serious hardware problems
      \newline $\Rightarrow$ TrueTime as trustworthy as any other piece of software
      }
    \end{itemize}
    \pause
    \begin{figure}
      \centering
      \includegraphics[width=.6\textwidth]{figures/epsilon.png}
      \caption{Distribution of TrueTime $\epsilon$ values}
    \end{figure}
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\section{Conclusions \& Future Work}
\begin{frame}
  \frametitle{Conclusions \& Future Work}
  \begin{itemize}
    \item{5 years from Spanner's inception to iterate to the current design}
    \item{Combines \& extends on ideas from database \& systems community}
    \item{Linchpin: TrueTime}
    \item{Build distributed systems with much stronger time semantics}
    \item{No longer depend on loosely synchronized clocks \& weak time APIs}
    \pause
    \item{Future Work:}
    \begin{itemize}
      \item{Implement optimistical reads in parallel}
      \item{Support direct changes of Paxos configurations}
      \item{Improve single-node performance}
      \item{Support movement of processes between datacenters}
    \end{itemize}
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\begin{frame}
	\frametitle{Thank you for your attention!}
	\begin{center}
		\huge{Questions?}
	\end{center}
	%\par{References: \tiny{
	%	\def\newblock{\hskip .11em plus .33em minus .07em}
	%	\bibliography{\jobname}
	%	}
	%}
\end{frame}

\end{document}

### 2016-01-27 presentation/presentation_spanner.tex
Add visualization of RW transaction

@@
 
-\usetikzlibrary{arrows,positioning,calc}
+\usetikzlibrary{arrows,positioning,calc,shapes}
 
@@
 {
-Supervisor: Dipl.-Ing. Nikolaus Augsten, PhD \newline \newline \newline
-University of Salzburg \\ 
-\medskip
-Daniel.Kocher@stud.sbg.ac.at
+Summarizing Talk \newline
+Efficent Algorithms Seminar \newline
+University of Salzburg
 }
@@
       \item{Difficult to use for applications with complex, evolving schemas}
-      \item{Only \emph{eventually-consistent} (no strong consistency)}
+      \item{Only \textbf{eventually-consistent} (no strong consistency)}
     \end{itemize}
@@
     \item{Each version is automatically timestamped with its commit time}
-    \item{Garbage Collection \& Read at old timestamps}
+    \item{Garbage collection \& reads at old timestamps}
     \item{\textbf{General-purpose transactions}}
@@
       }
-      \item{Timestamp are assigned to data $\Rightarrow$
+      \item{Timestamps are assigned to data $\Rightarrow$
         \textbf{multi-version database}
       }
-      \item{Tablet states are stored on Colossus (successor of the GFS)}
-      \item{Single Paxos state machine on top of each tablet to $\Rightarrow$
+      \item{Tablet states are stored on Colossus}
+      \item{Single Paxos state machine on top of each tablet $\Rightarrow$
         \textbf{Replication}
@@
       \end{itemize}
-      \item{State of each Transaction manager is stored in Paxos group}
+      \item{State of each transaction manager is stored in Paxos group}
     \end{itemize}
@@
   \frametitle{Directories \& Placement}
-  \only<1-4>{
+  \only<1-3>{
     \begin{itemize}
@@
         \item{Contain set of contiguous keys sharing common prefix}
-        \item{Prefix allows to control locality of data}
+        \item{Control locality of data}
         \item{Unit of data placement (same replication configuration)}
@@
       \end{enumerate}
-      \pause
       \item{\textbf{Movedir:} background task to move directories}
@@
   }
-  \only<5>{
+  \only<4>{
     \begin{figure}
@@
   \only<2>{
+    \begin{itemize}
+      \item{Databases must be partitioned in one or more hierarchies}
+      \item{\textbf{Directory table}: table at the top}
+      \item{Directory contains ordered rows starting at a directory table key}
+      \item{Interleaving tables to form directories is significant}
+      \item{Locality relationships between multiple tables $\Rightarrow$ good
+        performance
+      }
+    \end{itemize}
+
     \begin{figure}
@@
     \end{figure}
-    \begin{itemize}
-      \item{Interleaving of tables to form directories is significant}
-      \item{Locality relationships between multiple tables $\Rightarrow$ good
-        performance
-      }
-    \end{itemize}
   }
@@
   \only<2>{
+    \begin{itemize}
+      \item{Time references: GPS and atomic clocks}
+      \item{Set of \textbf{time master} machines per datacenter}
+      \item{\textbf{Armageddon masters}: masters using atomic clocks}
+      \item{\textbf{Timeslave daemon} on each machine}
+      \pause
+      \item{Masters compare their time references regularly}
+      \item{Daemons poll a variety of masters (liar detection)}
+      \pause
+      \item{$\epsilon$ is the instantaneous error bound (typically 1 - 7ms)}
+      \item{$\overline{\epsilon}$ is the average error bound (typically 4ms)}
+      \item{$\epsilon$ is derived from applied worst-case local clock drift ($200\mu s/sec$)}
+    \end{itemize}
+  }
+  \only<3>{
     \begin{figure}
@@
 
-        \draw[blue!35, fill=blue!35] (left-bound-top.center) rectangle (right-bound-bottom.center);
+        \draw[blue!35, fill=blue!35] (left-bound-top.center) rectangle
+          (right-bound-bottom.center);
         \node (two-e-invis) at ($(left-bound.south)!0.5!(right-bound.south)$) {};
@@
         \draw[<->] (left-bound-e.center) -- (right-bound-e.center);
-        %\node[below right= 1 and 2.5 of left] {$2\epsilon$};
       \end{tikzpicture}
@@
     \end{figure}
-  }
-  \only<3>{
-    \begin{itemize}
-      \item{Time references: GPS and atomic clocks (different failure modes)}
-      \item{Set of \textbf{time master} machines per datacenter}
-      \item{\textbf{Armageddon masters}: masters using atomic clocks}
-      \item{\textbf{timeslave daemon} per machine}
-      \pause
-      \item{Masters compare their time references regularly}
-      \item{Daemons poll a variety of masters (liar detection)}
-      \pause
-      \item{$\epsilon$ is the instantaneous error bound (typically 1 - 7ms)}
-      \item{$\overline{\epsilon}$ is the average error bound (typically 4ms)}
-      \item{$\epsilon$ is derived from applied worst-case local clock drift ($200\mu s/sec$)}
-    \end{itemize}
   }
@@
       \item{must predeclare to not include writes}
-      \item{can proceed on any sufficiently up-to-date-replica (also snapshot reads)}
+      \item{can proceed on any sufficiently up-to-date replica (also snapshot
+        reads)
+      }
     \end{itemize}
@@
   \frametitle{Timestamps in RW Transactions}
+  \only<1>{
     \begin{itemize}
@@
     \end{itemize}
+  }
+  \only<2>{
+    \begin{itemize}
+      \item{Let $e^{server}_i$ denote the arrival event of commit request for
+        $T_i$
+      }
+      \item{Two rules guarantee the external consistency invariant:}
+      \begin{enumerate}
+        \item{\textbf{Start} \newline
+          Coordinator leader assigns commit timestamp \newline
+          $s_i \geq TT.now().latest$ after $e^{server}_i$
+        }
+        \item{\textbf{Commit Wait} \newline
+          Coordinator leader ensures clients cannot see \newline
+          data commited by $T_i$ until $TT.after(s_i)$ is true
+        }
+      \end{enumerate}
+    \end{itemize}
+  }
 \end{frame}
@@
   \frametitle{Read-Write Transactions}
+  \only<1-3>{
     \begin{itemize}
@@
     \end{itemize}
+  }
+  \only<4-9>{
+    \begin{figure}
+      \centering
+      \begin{tikzpicture}[thick]
+        \node (left-end-t1) at (0,2) {};
+        \node[right = 9 of left-end-t1] (right-end-t1) {};
+        \draw[ultra thick, ->, >=stealth] (left-end-t1.center) -- (right-end-t1.center)
+          node[right] (right-end-t1-t) {$t$};
+          \node[left = 0 of left-end-t1.west] (left-end-t1-t) {$\mathbf{T_C}$};
+
+        \node[below = 1 of left-end-t1] (left-end-t2) {};
+        \node[right = 9 of left-end-t2] (right-end-t2) {};
+        \draw[->, >=stealth] (left-end-t2.center) -- (right-end-t2.center)
+          node[right] (right-end-t2-t) {$t$};
+        \node[left = 0 of left-end-t2.west] (left-end-t2-t) {$T_2$};
+
+        \node[below = 1 of left-end-t2] (left-end-t3) {};
+        \node[right = 9 of left-end-t3] (right-end-t3) {};
+        \draw[->, >=stealth] (left-end-t3.center) -- (right-end-t3.center)
+          node[right] (right-end-t3-t) {$t$};
+        \node[left = 0 of left-end-t3.west] (left-end-t3-t) {$T_3$};
+
+        \node[right = 0.25 of left-end-t1] (acquire-locks-t1) {};
+        \node[right = 1.05 of left-end-t2] (acquire-locks-t2) {};
+        \node[right = 0.65 of left-end-t3] (acquire-locks-t3) {};
+        \foreach \x in {1, 2, 3}{
+          \draw[blue] (acquire-locks-t\x.center) -- ++(0, -0.25);
+          \draw[blue] (acquire-locks-t\x.center) -- ++(0, 0.25);
+        }
+
+        \node[above = 0.75 of acquire-locks-t1]
+          (acquire-locks-t) {\scriptsize\color{blue}Acquire all locks};
+        \
+
+        \only<5-9>{
+          \foreach \x in {1, 2, 3}{
+            \node[right = -0.1 of acquire-locks-t\x] (compute-s-t\x) {};
+            %\draw (compute-s-t\x.center) -- ++(0, -0.15);
+            %\draw (compute-s-t\x.center) -- ++(0, 0.15);
+          }
+          \node[below = 1 of acquire-locks-t3.center]
+            (compute-s-t) {\scriptsize $s=TT.now()$};
+          \draw[<-, >=stealth, dotted] (compute-s-t1.center) -- ++(0, -3.5);
+          \draw[<-, >=stealth, dotted] (compute-s-t2.center) -- ++(0, -2.25);
+          \draw[<-, >=stealth, dotted] (compute-s-t3.center) -- ++(0, -0.9);
+        }
+        \only<6-9>{
+          \node[right = 1 of compute-s-t2] (start-log-t2) {};
+          \node[right = 1.75 of compute-s-t3] (start-log-t3) {};
+          \node[above = 2 of start-log-t2] (start-log-t)
+            {\scriptsize Start Logging};
+          \draw[<-, >=stealth, dotted] (start-log-t2.center) -- ++(0, 2.25);
+          \draw[<-, >=stealth, dotted] (start-log-t3.center) -- ++(0, 3.5);
+        }
+        \only<7-9>{
+          \node[right = 1 of start-log-t3] (end-log-t3) {};
+          \node[right = 1.75 of start-log-t2] (end-log-t2) {};
+          \node[above = 2 of end-log-t2] (end-log-t)
+            {\scriptsize Logging done};
+          \draw[<-, >=stealth, dotted] (end-log-t2.center) -- ++(0, 2.25);
+          \draw[<-, >=stealth, dotted] (end-log-t3.center) -- ++(0, 3.5);
+          \node[right = 5.5 of compute-s-t1] (compute-s-global) {};
+          \draw[<-, >=stealth, black!50!green] (compute-s-global) ++(-0.1, 0) --
+            (end-log-t3.north east);
+          \draw[<-, >=stealth, black!50!green] (compute-s-global) ++(-0.65, 0) --
+            (end-log-t2.north east);
+          \draw[black!50!green, ultra thick] (compute-s-global.center) -- ++(0, -0.25);
+          \draw[black!50!green, ultra thick] (compute-s-global.center) -- ++(0, 0.25);
+
+          \node[above = 0.25 of compute-s-global] (compute-s-global-t)
+            {\scriptsize\color{black!50!green} Overall $s$};
+          \node[below = 0 of end-log-t3] (prepared-t)
+            {\scriptsize\color{black!50!green}Prepared; Send $s$};
+        }
+        \only<8-9>{
+          \node[right = 0.5 of compute-s-global] (commit-wait-done) {};
+          \node[below = 3.45 of commit-wait-done] (commit-wait-done-t)
+            {\scriptsize Commit Wait done};
+          \draw[<-, >=stealth, dotted] (commit-wait-done.center) -- ++(0, -3.5);
+          \node[right = 0 of commit-wait-done] (release-locks-t1) {};
+          \draw[red] (release-locks-t1.center) -- ++(0, -0.25);
+          \draw[red] (release-locks-t1.center) -- ++(0, 0.25);
+          \node[above right = 0.75 and -0.75 of release-locks-t1]
+            (release-locks-t1-t) {\scriptsize\color{red}Release all locks};
+        }
+        \only<9>{
+          \node[right = -0.15 of release-locks-t1] (notify-t1) {};
+          \node[right = 3.1 of end-log-t2] (notify-t2) {};
+          \node[right = 3.35 of end-log-t3] (notify-t3) {};
+          \foreach \x in {2, 3}{
+            \draw[->, >=stealth, black!50!green] (notify-t1.center) --
+              (notify-t\x.center); 
+          }
+          %\node[below = 0 of notify-t3] (notify-t)
+          %  {\scriptsize\color{black!50!green}Notify participants};
+          \node[right = -0.15 of notify-t2] (release-locks-t2) {};
+          \node[right = -0.15 of notify-t3] (release-locks-t3) {};
+          \draw[red] (release-locks-t2.center) -- ++(0, -0.25);
+          \draw[red] (release-locks-t2.center) -- ++(0, 0.25);
+          \draw[red] (release-locks-t3.center) -- ++(0, -0.25);
+          \draw[red] (release-locks-t3.center) -- ++(0, 0.25);
+        }
+     \end{tikzpicture}
+      \caption{RW transaction illustration}
+    \end{figure}
+  }
 \end{frame}

### 2016-01-31 presentation/presentation_spanner.tex
Final version

@@
 
-\title[]{Spanner: Google's Globally-Distributed Database}
+\title[]{Spanner: Google's Globally-Distributed Database*}
 \subtitle[]{}
@@
 \titlepage
+{\footnotesize* James C. Corbett et al. (26 authors), OSDI 2012.}
 \end{frame}
@@
     \item{High availability}
-    \item{Two "competitors": \emph{Bigtable} \& \emph{Megastore}}
     \pause
@@
     \pause
-    \item{Why not use Megastore}
+    \item{Why not use Megastore?}
     \begin{itemize}
-      \item{Poor write throughput}
+      \item{Poor performance}
     \end{itemize}
-
   \end{itemize}
@@
   \frametitle{Fundamentals}
+  \only<1>{
     \begin{itemize}
-    \item{Transaction $i$ is denoted $T_i$}
+      %\item{Transaction $i$ is denoted $T_i$}
       %\item{RW \ldots Read-Write (transaction)}
       %\item{RO \ldots Read-Only (transaction)}
-    \item{\textbf{Externally-consistent:} \newline
-      If $T_1$ preceeds $T_2$ (Commit of $T_1$ happens before Begin of $T_2$, no
-      overlap), then $T_1$ is serialized first
+      \item{\textbf{External consistency:} \newline
+        If $T_1$ preceeds $T_2$, then $T_1$ is serialized first
       }
-    \pause
+      \item{\textbf{Paxos (Replication):}
+        \begin{itemize}
+          \item{Solves problem of resilient replication of data}
+          \item{Data eventually propagates to all nodes}
+          \item{Different nodes always see the same data}
+          \item{Majority of nodes up $\Rightarrow$ Writes/Reads processed
+            correctly
+          }
+          \item{Single node is elected as leader and initiates consensus}
+          \item{Guarantees \textbf{consistency}}
+        \end{itemize}
+      }
+    \end{itemize}
+  }
+  \only<2>{
+    \begin{itemize}
       \item{\textbf{Two-Phase Locking (2PL):} \newline
@@
     \end{itemize}
+  }
 \end{frame}
@@
   transactions \newline
-  \pause
   \begin{itemize}
-    \item{Timestamps reflect serialization order (linearizability)}
-    \item{Key enabler: novel TrueTime API exposing clock uncertainty}
+    \item{Timestamps reflect serialization order}
+    \item{Key enabler: novel \textbf{TrueTime} API \textbf{exposing clock uncertainty}}
   \end{itemize}
@@
 \begin{frame}
-  \frametitle{Implementation (Top-down)}
-  \only<1-3>{
-    \begin{itemize}
-      \item{Spanner deployment is called \textbf{universe}}
-      \item{Organized as a set of \textbf{zones}}
-      \item{Zones are the unit of administrative deployment \& physical isolation}
-      \item{Set of zones is the set of locations for replication}
-      \pause
-      \item{Each zone consists of}
-      \begin{itemize}
-        \item{several \textbf{spanservers} to store the data}
-        \item{a \textbf{zonemaster} to assign data to spanservers}
-        \item{a \textbf{location proxy} for clients to locate the assigned
-          spanservers
-        }
-      \end{itemize}
-      \pause
-      \item{\textbf{universe master:} console to display status information about zones}
-      \item{\textbf{placement driver:}}
-      \begin{itemize}
-        \item{handles automated data movement across zones}
-        \item{moves data due to updated replication constraints or load balancing}
-      \end{itemize}
-    \end{itemize}
-  }
-  \only<4>{
+  \frametitle{Implementation}
+  %\only<1-3>{
+  %  \begin{itemize}
+  %    \item{Spanner deployment is called \textbf{universe}}
+  %    \item{Organized as a set of \textbf{zones}}
+  %    \item{Zones are the unit of administrative deployment \& physical isolation}
+  %    \item{Set of zones is the set of locations for replication}
+  %    \pause
+  %    \item{Each zone consists of}
+  %    \begin{itemize}
+  %      \item{several \textbf{spanservers} to store the data}
+  %      \item{\textbf{location proxies} for clients to locate the assigned
+  %        spanservers
+  %      }
+  %      \item{a \textbf{zonemaster} to assign data to spanservers}
+  %    \end{itemize}
+  %    \pause
+  %    \item{\textbf{universe master:} console to display status information about zones}
+  %    \item{\textbf{placement driver:}}
+  %    \begin{itemize}
+  %      \item{handles automated data movement across zones}
+  %      \item{moves data due to updated replication constraints or load balancing}
+  %    \end{itemize}
+  %  \end{itemize}
+  %}
+  %\only<4>{
     \begin{figure}
@@
     \end{figure}
-  }
+  %}
 \end{frame}
@@
       \begin{itemize}
-        \item{a \textbf{lock table} for concurrency control: \texttt{(key range) $\rightarrow$ lock state}}
+        \item{a \textbf{lock table} for concurrency control: state for 2PL}
         \item{a \textbf{transaction manager} (TM) for distributed transactions}
@@
       \begin{itemize}
-        \item{One participant group is chosen as \textbf{coordinator}}
+        \item{One participant group is chosen as \textbf{coordinator group}}
         \item{Participant leader of this group: \textbf{coordinator leader}}
@@
   \frametitle{Directories \& Placement}
-  \only<1-3>{
+  \only<1-2>{
     \begin{itemize}
@@
       \begin{itemize}
-        \item{Bucketing abstraction on top of key-value mappings}
+        \item{Bucketing abstraction on top of tablet}
         \item{Contain set of contiguous keys sharing common prefix}
-        \item{Control locality of data}
+        \item{Control locality of data (pyhsical)}
         \item{Unit of data placement (same replication configuration)}
-        \item{Smallest unit to specify geographic-replication properties (\textbf{placement})}
+        \item{Smallest unit to specify geographic-replication properties
+          (\textbf{placement})
+        }
       \end{itemize}
@@
       \item{Data is moved directory-wise between Paxos groups}
-      \item{Tablets may have multiple \textbf{partitions} of row space \newline
-        $\Rightarrow$ co-locate directories frequently accessed together
+      \item{Tablets may have multiple directories \newline
+        $\Rightarrow$ co-locate data frequently accessed together
       }
-      \pause
       \item{Placements can be controlled in two dimensions:}
@@
       \end{enumerate}
-      \item{\textbf{Movedir:} background task to move directories}
+      % \item{\textbf{Movedir:} background task to move directories}
     \end{itemize}
   }
-  \only<4>{
+  \only<3>{
     \begin{figure}
@@
     \begin{itemize}
-      \item{Features:}
-      \begin{itemize}
-        \item{Schematized semi-relational tables}
-        \item{SQL-like query language}
-        \item{General-purpose transactions}
-      \end{itemize}
+      %\item{Features:}
+      %\begin{itemize}
+      %  \item{Schematized semi-relational tables}
+      %  \item{SQL-like query language}
+      %  \item{General-purpose transactions}
+      %\end{itemize}
       \item{Layered on top of directory-bucketed key-value mappings}
-      \item{Applications create \textbf{databases} in universe}
-      \item{Database may contain unlimited number of schematized \textbf{tables}}
-      \item{Table contains rows (must have names), columns \& versioned values}
+      \item{Applications create \textbf{databases}}
+      \item{Database may contain any number of schematized \textbf{tables}}
+      \item{Table contains rows, columns \& versioned values}
     \end{itemize}
@@
       \item{\textbf{Directory table}: table at the top}
-      \item{Directory contains ordered rows starting at a directory table key}
-      \item{Interleaving tables to form directories is significant}
-      \item{Locality relationships between multiple tables $\Rightarrow$ good
-        performance
+      \item{Directory contains hierarchically ordered rows}
+      \item{Interleaving tables $\Rightarrow$ locality relationships of multiple
+        tables \newline$\Rightarrow$ significant for performance
       }
@@
   \only<1>{
-    \begin{table}
-      \centering
+    \begin{center}
+      {\footnotesize
       \begin{tabular}{|l||l|}
@@
       \end{tabular}
-      \caption{TrueTime API ($t$ is of type $TTstamp$)}
-    \end{table}
+      }
+    %  \caption{TrueTime API ($t$ is of type $TTstamp$)}
+    \end{center}
     Time is represented as a $TTinterval$ (interval with bounded uncertainty)
     \newline\newline
-    Let $e$ be an event and $t_{abs}(e)$ denote the absolute time of $e$.
-    \newline
-    For $tt = TT.now()$, the following property holds:
-    \begin{center}
-      $tt.earliest \leq t_{abs}(e) \leq tt.latest$
-    \end{center}
+    %Let $e$ be an event and $t_{abs}(e)$ denote the absolute time of $e$.
+    %\newline
+    %For $tt = TT.now()$, the following property holds:
+    %\begin{center}
+    %  $tt.earliest \leq t_{abs}(e) \leq tt.latest$
+    %\end{center}
+    \textbf{Guarantee:} $TT.now().earliest \leq t_{abs}(e) \leq TT.now().latest$
   }
@@
       \item{Set of \textbf{time master} machines per datacenter}
-      \item{\textbf{Armageddon masters}: masters using atomic clocks}
+      %\item{\textbf{Armageddon masters}: masters using atomic clocks}
       \item{\textbf{Timeslave daemon} on each machine}
-      \pause
-      \item{Masters compare their time references regularly}
-      \item{Daemons poll a variety of masters (liar detection)}
-      \pause
-      \item{$\epsilon$ is the instantaneous error bound (typically 1 - 7ms)}
+      \item{Masters compare time references regularly}
+      \item{Daemons poll a variety of masters}
+      \item{$\epsilon$ is the \textbf{instantaneous error bound} (typically 1 -
+        7ms)
+      }
       \item{$\overline{\epsilon}$ is the average error bound (typically 4ms)}
@@
   \begin{itemize}
-    \item{TrueTime is used to guarantee correctness properties}
-    \item{Properties are used to implement}
+    \item{TrueTime used to guarantee correctness properties to implement:}
     \begin{itemize}
@@
       \item{Lock-free read-only transactions}
-      \item{Non-blocking reads in the past}
+      \item{Non-blocking reads in the past (snapshot reads)}
     \end{itemize}
@@
   \only<1>{
-    \footnotesize
-    \begin{table}
+    \begin{center}
+      {\footnotesize
       \begin{tabular}{|l||l|l|}
@@
       \end{tabular}
-      \caption{Supported types of reads \& writes}
-    \end{table}
+      }
+    %  \caption{Supported types of reads \& writes}
+    \end{center}
   }
   \begin{itemize}
-    \item{Standalone write $\Rightarrow$ RW transaction}
-    \item{Non-snapshot standalone read $\Rightarrow$ RO transaction}
-    \item{RO transactions}
-    \begin{itemize}
-      \item{must predeclare to not include writes}
-      \item{can proceed on any sufficiently up-to-date replica (also snapshot
-        reads)
+    \item{Standalone write $\Rightarrow$ Read-Write (RW) transaction}
+    \item{Non-snapshot standalone read $\Rightarrow$ Read-Only (RO) transaction}
+    \item{RO transactions must predeclare to not include writes}
+    \item{RO transactions \& snapshot reads proceed on any
+      \textbf{sufficiently up-to-date} replica
     }
-    \end{itemize}
   \end{itemize}
@@
     }
-    \item{Leaders must not abdicate before $TT.after(s_{max})$, where $s_{max}$
-      denotes the maximum timestamp used by a leader
-    }
+    \item{Leaders must not abdicate before $TT.after(s_{max})$}
+    \item{$s_{max}$ \ldots maximum timestamp used by a leader}
   \end{itemize}
@@
 %-------------------------------------------------------------------------------
-\subsubsection{Timestamps in RW Transactions}
+\subsubsection{Read-Write Transactions}
 \begin{frame}
-  \frametitle{Timestamps in RW Transactions}
+  \frametitle{Read-Write Transactions}
   \only<1>{
     \begin{itemize}
-      \item{Timestamp assignment only in between the two phases (2PL)}
-      \item{\textbf{Monotonicity variant:} \newline
+      \item{Uses 2PL: Timestamp assignment in between the two phases}
+      \item{\textbf{Monotonicity invariant:} \newline
         Within each Paxos group, Spanner assigns timestamps to Paxos writes in
@@
       \item{Enforced across leaders by disjointness invariant: \newline
-        Leader must only assign timestamps within the interval of leader lease
+        Leader must only assign timestamps within interval of leader lease
       }
@@
     \begin{itemize}
+      \item{Writes buffered at client until commit} 
+      \item{Client has completed all reads \& buffered all writes $\Rightarrow$ 2PC}
       \item{Let $e^{server}_i$ denote the arrival event of commit request for
@@
           data commited by $T_i$ until $TT.after(s_i)$ is true
+
         }
       \end{enumerate}
+      %\pause
+      %\item{Non-coordinator-participant leader}
+      %\begin{enumerate}
+      %  \item{Acquires write locks}
+      %  \item{Chooses prepare timestamp according to monotonicity}
+      %  \item{Logs prepare record through Paxos}
+      %  \item{Notifies coordinator of its prepare timestamp}
+      %\end{enumerate}
+      %\pause
+      %\item{Coordinator leader}
+      %\begin{enumerate}
+      %  \item{Aquires write locks}
+      %  \item{Chooses timestamp for whole transaction}
+      %  \item{Commit timestamp $s$ must be}
+      %  \begin{itemize}
+      %    \item{$\geq$ all prepare timestamps}
+      %    \item{$> TT.now().latest$ at the time the commits were received}
+      %    \item{$>$ any previously assigned timestamp}
+      %  \end{itemize}
+      %  \item{Logs Commit record through Paxos}
+      %\end{enumerate}
     \end{itemize}
   }
-\end{frame}
-
-%-------------------------------------------------------------------------------
-\subsubsection{Serving Reads at a Timestamp}
-\begin{frame}
-  \frametitle{Serving Reads at a Timestamp}
-  \begin{itemize}
-    \item{Monotonicity invariant allows to determine if a replica's state is
-      \textbf{sufficiently up-to-date} to satisfy a read
-    }
-    \item{\textbf{Safe time} $t_{safe}$ at each replica}
-    \item{Replica satisfies read if $t \leq t_{safe}$}
-    \item{$t_{safe} = min\left(t^{Paxos}_{safe}, t^{TM}_{safe}\right)$}
-  \end{itemize}
-\end{frame}
-
-%-------------------------------------------------------------------------------
-\subsubsection{Timestamps in RO Transactions}
-\begin{frame}
-  \frametitle{Timestamps in RO Transactions}
-  \begin{itemize}
-    \item{Two phases:}
-    \begin{enumerate}
-      \item{Assign timestamp $s_{read}$}
-      \item{Execute transaction's reads as snapshot reads at $s_{read}$}
-    \end{enumerate}
-    \item{Spanner assigns the oldest timestamp that preserves external consistency
-      to reduce the changes of blocking
-    }
-  \end{itemize}
-\end{frame}
-
-%-------------------------------------------------------------------------------
-\subsection{Details}
-
-\subsubsection{Read-Write Transactions}
-\begin{frame}
-  \frametitle{Read-Write Transactions}
-  \only<1-3>{
-    \begin{itemize}
-      \item{RW transactions buffered at client until Commit}
-      \item{Reads do not see effects of transaction's writes}
-      \item{Client has completed all reads \& buffered all writes $\Rightarrow$ 2PC}
-      \pause
-      \item{Non-coordinator-participant leader}
-      \begin{enumerate}
-        %\item{Acquires write locks}
-        \item{Chooses prepare timestamp according to monotonicity}
-        \item{Logs prepare record through Paxos}
-        \item{Notifies coordinator of its prepare timestamp}
-      \end{enumerate}
-      \pause
-      \item{Coordinator leader}
-      \begin{enumerate}
-        %\item{Aquires write locks}
-        \item{Chooses timestamp for whole transaction}
-        \item{Commit timestamp $s$ must be}
-        \begin{itemize}
-          \item{$\geq$ all prepare timestamps}
-          \item{$> TT.now().latest$ at the time the Commits were received}
-          \item{$>$ any previously assigned timestamp}
-        \end{itemize}
-        \item{Logs Commit record through Paxos}
-      \end{enumerate}
-    \end{itemize}
-  }
-  \only<4-9>{
+  \only<3-8>{
     \begin{figure}
@@
           node[right] (right-end-t1-t) {$t$};
-          \node[left = 0 of left-end-t1.west] (left-end-t1-t) {$\mathbf{T_C}$};
+          \node[left = 0 of left-end-t1.west] (left-end-t1-t) {$\mathbf{P_{CL}}$};
 
@@
           node[right] (right-end-t2-t) {$t$};
-        \node[left = 0 of left-end-t2.west] (left-end-t2-t) {$T_2$};
+        \node[left = 0 of left-end-t2.west] (left-end-t2-t) {$P_2$};
 
@@
           node[right] (right-end-t3-t) {$t$};
-        \node[left = 0 of left-end-t3.west] (left-end-t3-t) {$T_3$};
+        \node[left = 0 of left-end-t3.west] (left-end-t3-t) {$P_3$};
 
@@
 
-        \only<5-9>{
+        \only<4-8>{
           \foreach \x in {1, 2, 3}{
@@
         }
-        \only<6-9>{
+        \only<5-8>{
           \node[right = 1 of compute-s-t2] (start-log-t2) {};
@@
         }
-        \only<7-9>{
+        \only<6-8>{
           \node[right = 1 of start-log-t3] (end-log-t3) {};
@@
         }
-        \only<8-9>{
+        \only<7-8>{
           \node[right = 0.5 of compute-s-global] (commit-wait-done) {};
@@
         }
-        \only<9>{
+        \only<8>{
           \node[right = -0.15 of release-locks-t1] (notify-t1) {};
@@
   }
+  \only<9>{
+    \begin{center}
+      {\footnotesize
+      \begin{tabular}{|l||l|l|}
+        \hline
+        Operation & CC & Replica required\tabularnewline
+        \hline\hline
+        Read-Write transaction & Pessimistic & leader\tabularnewline
+        \hline
+      \end{tabular}
+      }
+    \end{center}
+  }
+\end{frame}
+
+%-------------------------------------------------------------------------------
+\subsubsection{Serving Reads at a Timestamp}
+\begin{frame}
+  \frametitle{Serving Reads at a Timestamp}
+  \begin{center}
+    {\footnotesize
+    \begin{tabular}{|l||l|l|}
+      \hline
+      Operation & CC & Replica required\tabularnewline
+      \hline\hline
+      Snapshot Read, client-provided timestamp & Lock-free & any\tabularnewline
+      \hline
+      Snapshot Read, client-provided bound & Lock-free & any \tabularnewline
+      \hline
+    \end{tabular}
+    }
+  \end{center}
+  \begin{itemize}
+    \item{Monotonicity invariant allows to determine if a replica's state is
+      \textbf{sufficiently up-to-date} to satisfy a read
+    }
+    \item{\textbf{Safe time} $t_{safe}$ at each replica}
+    \item{Replica satisfies read at timestamp $t$ if $t \leq t_{safe}$}
+    \item{Performance benefits}
+    %\item{$t_{safe} = min\left(t^{Paxos}_{safe}, t^{TM}_{safe}\right)$}
+    %\item{$t^{Paxos}_{safe}$ = highest-applied Paxos write}
+    %\item{Transaction in between phases of 2PC
+    %  $\Rightarrow t^{TM}_{safe} = \infty$ \newline
+    %  Participant replica does not know if transaction will commit
+    %}
+  \end{itemize}
 \end{frame}
@@
   \frametitle{Read-Only Transactions}
+  \only<1>{
+    \begin{center}
+      {\footnotesize
+      \begin{tabular}{|l||l|l|}
+        \hline
+        Operation & CC & Replica required\tabularnewline
+        \hline\hline
+        Read-Only transaction & Lock-free & leader; any
+        \tabularnewline
+        \hline
+      \end{tabular}
+      }
+    %  \caption{Supported types of reads \& writes}
+    \end{center}
     \begin{itemize}
+      \item{Two phases:}
+      \begin{enumerate}
+        \item{Assign timestamp $s_{read}$}
+        \item{Execute transaction's reads as snapshot reads at $s_{read}$}
+      \end{enumerate}
+      \item{Spanner assigns $s_{read}$ oldest timestamp to preserves external
+        consistency
+      }
+    \end{itemize}
+  }
+  \only<2>{
+    \begin{itemize}
+      \item{Negotiation phase between all Paxos groups involved}
       \item{\textbf{scope} expression summarizing the keys that will be read}
-    \item{Served by single Paxos group \newline
-      $\Rightarrow$ client issues transaction to group leader
+      \item{Scope's values served by single Paxos group
+        \begin{itemize}
+          \item{Client issues RO transaction to group leader}
+          \item{Leader assigns $s_{read}$ = timestamp of last committed write}
+        \end{itemize}
       }
-    \item{Served by multiple Paxos groups \newline
-      $\Rightarrow$ client executes reads at $s_{read} = TT.now().latest$
+      \item{Scope's values served by multiple Paxos groups
+        \begin{itemize}
+          \item{Negotiation between leaders possible}
+          \item{Simpler: reads execute at $s_{read} = TT.now().latest$}
+        \end{itemize}
       }
@@
   \begin{itemize}
-    \item{TrueTime enables this feature}
-    \item{Use of standard transaction infeasible}
+    \item{Enabled by TrueTime}
+    %\item{Use of standard transaction infeasible (millions of participants)}
     \item{Non-blocking variant of standard transaction}
     \begin{enumerate}
-      \item{Explicit assignment of a future timestamp (Prepare phase)}
+      \item{Explicit assignment of a future timestamp}
       \item{Reads \& writes synchronize with any registered schema-change
@@
     \textbf{Latency [ms] (less is better)}
-    \begin{table}
-      \centering
+    \begin{center}
+      {\footnotesize
       \begin{tabular}{|l||l|l|l|}
@@
       \end{tabular}
-      \caption{Latency experiments}
-    \end{table}
-    \begin{itemize}
-      \item{Sufficiently few operations to avoid queuing}
-      \item{Increasing number of replicas:}
-      \begin{itemize}
-        \item{Latency stays roughly constant with less standard deviation}
-        \item{Latency to achieve quorum less sensitive to stragglers}
-      \end{itemize}
-    \end{itemize}
       }
-  \only<3>{
+    %  \caption{Latency experiments}
+    \end{center}
+    %\begin{itemize}
+    %  \item{Sufficiently few operations to avoid queuing}
+    %  \item{Increasing number of replicas:}
+    %  \begin{itemize}
+    %    \item{Latency stays roughly constant with less standard deviation}
+    %    \item{Latency to achieve quorum less sensitive to stragglers}
+    %  \end{itemize}
+    %\end{itemize}
     \textbf{Throughput [Kops/sec] (more is better)}
-    \begin{table}
-      \centering
+    \begin{center}
+      {\footnotesize
       \begin{tabular}{|l||l|l|l|}
@@
       \end{tabular}
-      \caption{Throughput experiments}
-    \end{table}
-    \begin{itemize}
-      \item{Sufficiently many operations to saturate servers' CPUs}
-      \item{Increasing number of replicas:}
-      \begin{itemize}
-        \item{Snapshot read throughput increases almost linear}
-      \end{itemize}
-    \end{itemize}
+      }
+    %  \caption{Throughput experiments}
+    \end{center}
+    %\begin{itemize}
+    %  \item{Sufficiently many operations to saturate servers' CPUs}
+    % \item{Increasing number of replicas:}
+    %  \begin{itemize}
+    %    \item{Snapshot read throughput increases almost linear}
+    %  \end{itemize}
+    %\end{itemize}
   }
@@
   \frametitle{Availability}
-  \only<1>{
   \begin{itemize}
@@
       \item{Universe with 5 zones $Z_i$, each of which has 25 spanservers}
-        \item{Database sharded into 1250 Paxos groups}
-        \item{100 clients constantly issued snapshot reads (50K reads/sec)}
-        \item{All leaders explicitly place in $Z_1$}
+      \item{All leaders explicitly placed in $Z_1$}
       \item{5s into each test, all servers in one zone were killed}
@@
   \end{itemize}
-  }
-  \only<2>{
   \begin{figure}
@@
   \end{figure}
-    \begin{itemize}
-      \item{Availability benefits of running Spanner in multiple datacenters}
-      %\item{Killing $Z_2$ (non-leader): no effect on read throughput}
-      %\item{Killing $Z_1$ with notifications (leader-soft): minor effect}
-      %\item{Killing $Z_1$ without notifications (leafer-hard): severe effect}
-    \end{itemize}
-  }
 \end{frame}
@@
   \frametitle{TrueTime}
+  \begin{enumerate}
+    \item{Is $\epsilon$ truly a bound on clock uncertainty?
       \begin{itemize}
-    \item{Two questions:}
-    \begin{itemize}
-      \item{Is $\epsilon$ truly a bound on clock uncertainty?}
-      \item{How bad does $\epsilon$ get?}
-    \end{itemize}
-    \pause
-    \item{Answers:}
-    \begin{itemize}
+        \item{Local clock drifts $> 200\mu$s/sec would break assumptions}
         \item{Clock issues infrequent relative to more serious hardware problems
-      \newline $\Rightarrow$ TrueTime as trustworthy as any other piece of software
+          \newline $\Rightarrow$ TrueTime as trustworthy as any other piece of
+          software
         }
       \end{itemize}
+    }
     \pause
+    \item{How bad does $\epsilon$ get?}
+  \end{enumerate}
   \begin{figure}
@@
   \end{figure}
-  \end{itemize}
 \end{frame}
@@
     \item{5 years from Spanner's inception to iterate to the current design}
-    \item{Combines \& extends on ideas from database \& systems community}
+    %\item{Combines \& extends on ideas from database \& systems community}
     \item{Linchpin: TrueTime}
@@
     \begin{itemize}
-      \item{Implement optimistical reads in parallel}
+      \item{Support parallel reads}
       \item{Support direct changes of Paxos configurations}
-      \item{Improve single-node performance}
+      \item{Improve single-node performance for complex queries}
       \item{Support movement of processes between datacenters}

### 2016-02-16 paper/paper.tex
Add first version of paper

\documentclass[onecolumn, a4paper, 10pt]{article}

%%% Packages
\usepackage{graphicx}
\usepackage{balance}
\usepackage{tikz}
\usepackage{standalone}
\usepackage{colortbl}
\usepackage{subcaption}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{multirow,bigdelim}
\usepackage[pdftex, letterpaper, lmargin = 1.15in, rmargin = 1.15in]{geometry}

%%% TikZ libraries
\usetikzlibrary{backgrounds, positioning, decorations.pathreplacing, calc, fit}
\usetikzlibrary{shapes}

%%% Macros
\newcommand{\tbr}{{\color{red}\textbf{[TO BE REVISED]}}}
\newcommand{\tbd}{{\color{red}\textbf{[TO BE DONE]}}}

%\setlength{\parskip}{1em}

\begin{document}

% Title
\title{
  Spanner: Google's Globally-Distributed Database \newline
  Seminar Report
}

\author{
  Daniel Kocher \\
  Department of Computer Sciences \\
  University of Salzburg \\                                                    
  Daniel.Kocher@stud.sbg.ac.at
}

\maketitle

\section{Introduction \tbr}
\label{sec:introduction}

In the original paper~\cite{Corbett:2012}, Google introduced a
globally-distributed database referred to as \emph{Spanner}. This system is based
upon Paxos~\cite{Lamport:1998} state machines on machines spread across multiple
datacenters across the world. This global replication guarantees high availability
and also implies some desirable locality properties which benefit the performance.

Google already has two other highly scalable distributed databases, i.e.
Bigtable~\cite{Chang:2008} and Megastore~\cite{Baker:2011}. Hence, there must be
a reason, why they did not just use one of these but implemented another
distributed database. According to Google, it is difficult for applications/users
to deal with complex and evolving schemas. Moreover, Bigtable only guarantees to
be eventually-consistent. In contrast, Spanner provides much stronger consistency
guarantees. On the other hand, Megastore suffers from poor write throughput even
though it is used for many applications.

Therefore, Google decided to introduce another distributed, highly available
database system which evolved from a key-value store somehow similar to the one
used by Bigtable (but we will see that this is only partially correct). A core
feature provided by Spanner are so-called \emph{externally-consistent} distributed
general-purpose transactions. Of course, Spanner also provides an SQL-like query
language to enable applications/users to interact with Spanner in a well-known
way. The data is stored in schematized \emph{semi-relational} tables and the data
is versioned automatically. Google also implemented a garbage collection to deal
with old data. The replication can be dynamically configured in order to control
the locality of the replicas. This is useful to control write and read latency
as well as availability. Another main contribution of Spanner are
\emph{globally-consistent} reads at a specific timestamp. This and
externally-consistent reads and writes enable Spanner to provide a special type
of transaction, so-called \emph{schema-change} transactions, which provides atomic
schema changes without locking the whole database and all ongoing transactions.

To provide all these features, Google designed a novel \emph{TrueTime} API which
exposes clock uncertainty and enables Spanner to assign commit timestamps to
transactions which reflect serialization order.

\section{Implementation \tbr}
\label{sec:implementation}

A Spanner deployment is referred to as a \emph{universe} which consists of
multiple zones. A universe has two global components: A \emph{universemaster}
which, essentially, is a debugging console and shows the status of each zone. A
\emph{placement driver} which automatically moves data between zones due to
modified replication constraints or load balancing between zones.

Each zone is controlled by a \emph{zonemaster}. A zonemaster is responsible to
organize the data placement in a zone. The data is stored on so-called
\emph{spanservers} and the zonemaster assigns data to these spanservers. A zone
consists of multiple such spanservers which serve the data to clients. Spanner
somehow needs to enable the client to locate the spanservers which serve the data.
For this purpose, each zone contains multiple \emph{location proxies}.

According to Google, a zone can be seen as a Bigtable server deployment. The set
of zones also determines the locations across which the data can be replicated.
One can also think of a zone as geographically wide-spread locations, e.g.
continents. Using such a physical isolation, Spanner is resistent to events which
affect large areas, e.g. natural disasters or war. Hence, a well-configured
Spanner deployment is available almost all the time.
Figure~\ref{subfig:spanner-universe} illustrates the organization of a Spanner
universe.

\begin{figure}[ht]
  \centering
  \begin{subfigure}{0.45\textwidth}
    \includegraphics[width = \textwidth]{figs/spanner-server-organization.png}
    \caption{Spanner universe.~\cite{Corbett:2012}}
    \label{subfig:spanner-universe}
  \end{subfigure}
  \hfill
  \begin{subfigure}{0.45\textwidth}
    \includegraphics[width = \textwidth]{figs/spanserver-software-stack.png}
    \caption{Spanserver software stack.~\cite{Corbett:2012}}
    \label{subfig:spanserver-software-stack}
  \end{subfigure}
  \caption{Spanner hierarchy.}
  \label{fig:spanner-hierarchy}
\end{figure}

\subsection{Spanserver Software Stack \tbr}
\label{subsec:spanserver-software-stack}

Each spanserver also consists of multiple software layers (four at least). At the
bottom layer, every spanserver is responsible for between 100 and 1,000
\emph{tablets}. A tablet is bag of key-value mappings of the form
\begin{center}
  \texttt{(key:string, timestamp:int64)} $\rightarrow$ \texttt{value}
\end{center}
This is somehow similar to the tablet abstraction used in Bigtable. However,
Spanner \emph{automatically} assigns timestamps to data which makes Spanner a
\emph{multi-version} database. The state of a tablet is stored on a distributed
file system called \emph{Colossus}, the successor of the Google File
System~\cite{Ghemawat:2003}.

On top of each tablet, a Paxos state machine is implemented in order to support
consistent replication. One Paxos state machine is elect as leader. Leases are
assigned in a long-lived manner (default lease length is 10 seconds). A leader
implements three additional layers. A \emph{lock table} which stores the state
for two-phase locking and therefore implements the concurrency control. On top
of the lock table, a \emph{transaction manager} is implemented to coordinate
distributed transactions. In turn, this transaction manager is used to implement
a \emph{participant leader}. This leader is only utilized if a transaction
involves multiple Paxos groups (set of replicas). In this case, one of the
participant groups is chosen as coordinator and the leader of this group is
referred to as \emph{coordinator leader}.
Figure~\ref{subfig:spanserver-software-stack} shows the spanserver software stack
as described before.

\subsection{Directories \tbr}
\label{subsec:directories}

The next abstraction in the hierarchy is a bucketing abstraction. Google calls
this abstraction \emph{directory}. A directory is a set of contiguous keys.
All keys in a directory share a common prefix.

A directory is the unit of data placement/movement, hence only whole directories
are moved between Paxos groups, even across zones. Figure~\ref{subfig:directories}
illustrates this principle. Directory 2 is moved from Paxos group 1 to Paxos group
2 and at the same time from zone 2 to zone 3. Google shortly mentions that this
is an over-simplified view because a directory is split into multiple
\emph{fragments} if it grows too large. However, Google does not provide more
details.

All the data within a directory has the same replication configuration. As will
be described later, the concept of directories enables Google to improve the
performance of Spanner by placing directories in the same Paxos group which are
frequently accessed together. The fact that in general multiple directories are
placed in the same Paxos group is another difference between the key-value mapping
of Spanner and the one used in Bigtable. A Spanner tablet may contain multiple
contiguous ranges of keys (= directories) as opposed to a Bigtable tablet which
may only contain one contiguous range of keys. Figure~\ref{fig:comparison-tablets}
shows an example comparison of the two tablet types.

\begin{figure}[ht]
  \centering
  \begin{subfigure}{0.45\textwidth}
    \centering
    \begin{tabular}{|l||ll}
      \hline
      {\bfseries Key} & {\bfseries Value} \tabularnewline
      \hline\hline
      12341 & \ldots & \rdelim\}{3}{5mm}[Directory 1234] \tabularnewline
      12342 & \ldots \tabularnewline
      12343 & \ldots \tabularnewline
      \hline
      67891 & \ldots & \rdelim\}{4}{3mm}[Directory 6789] \tabularnewline
      67892 & \ldots \tabularnewline
      67893 & \ldots \tabularnewline
      67894 & \ldots \tabularnewline
      \hline
    \end{tabular}
    \caption{Spanner tablet}
    \label{subfig:spanner-tablet}
  \end{subfigure}
  \qquad
  \begin{subfigure}{0.45\textwidth}
    \centering
    \begin{tabular}{|l||l}
      \hline
      {\bfseries Key} & {\bfseries Value} \tabularnewline
      \hline\hline
      12341 & \ldots \tabularnewline
      12342 & \ldots \tabularnewline
      12343 & \ldots \tabularnewline
      12344 & \ldots \tabularnewline
      12345 & \ldots \tabularnewline
      12346 & \ldots \tabularnewline
      12347 & \ldots \tabularnewline
      \hline
    \end{tabular}
    \caption{Bigtable tablet}
    \label{subfig:bigtable-tablet}
  \end{subfigure}
  \caption{Comparison of Spanner's and Bigtable's tablets}
  \label{fig:comparison-tablets}
\end{figure}

This partitioning of the row space is critical for the performance of Spanner
because it allows Google to physically \emph{co-locate} data that is frequently
accessed together. Therefore, directories are the smallest unit for which an
application/user can specify geographic-replication properties. An application
can control replication in two dimension, i.e. the number and types of replicas,
and, the geographic placement of replicas.

\begin{figure}[ht]
  \centering
  \begin{subfigure}{0.45\textwidth}
    \includegraphics[width = \textwidth]{figs/directories}
    \caption{Directories (unit of data movement).~\cite{Corbett:2012}}
    \label{subfig:directories}
  \end{subfigure}
  \qquad
  \begin{subfigure}{0.45\textwidth}
    \includegraphics[width = \textwidth]{figs/interleaving-example.png}
    \caption{Two interleaved tables.~\cite{Corbett:2012}}
    \label{subfig:interleaved-tables}
  \end{subfigure}
  \caption{Directories and interleaved tables.}
  \label{fig:directories-and-tables}
\end{figure}

\subsection{Data Model \tbr}
\label{subsec:data-model}

There exists another layer on top of the directory-bucketed key-value mappings.
This layer is probably the most intuitive one for people who are used to common
database systems. The data model allows applications to create databases in a
universe each of which may contain any number of schematized
\emph{semi-relational} tables. A table has rows, columns and, as a third
dimension, versioned values. Spanner's tables are semi-relational because rows
must have names, i.e. every table must have an ordered set of one or more
primary-key columns. This is useful for applications because the locality of
data can be controlled by choosing keys accurately.

Therefore, Spanner requires the application/user to partition databases into
one or more hierarchies of tables. Based on this hierarchical information, Spanner
builds the directories. A directory contains hierarchically ordered rows from
all tables of a hierarchy which belong to the same key of the topmost table of
the hierarchy. Figure~\ref{subfig:interleaved-tables} shows an examples of this
concept. In this example, the \texttt{Users} table is the topmost table of the
hierarchy and the \texttt{Albums} table is one level below. As we can see, all
rows of \texttt{Albums} starting with \texttt{1} are located in the same directory
(\texttt{3365}). This directory has a common prefix, i.e. \texttt{1}, from the
\texttt{Users} table. The same principle can be seen for key \texttt{2} of the
\texttt{Users} table. This interleaving of tables exploits locality relationships
of multiple tables and therefore is significant for Spanner's performance.

\section{TrueTime \tbr}
\label{sec:truetime}

This section describes the key enabler of Spanner, i.e. the novel \emph{TrueTime}
API. Essentially, this API enables Google to provide timestamps for transactions
which reflect serialization order.

TrueTime represents time as a $TTinterval$. A $TTinterval$ is an interval with a
bounded time uncertainty and has two endpoints, i.e. $earliest$ and $latest$,
which are of type $TTstamp$ (which is a timestamp in the context of Spanner).
Table~\ref{tbl:truetime-api} summarizes the three methods provided by the TrueTime
API. However, the most important method is $TT.now()$ which returns a $TTinterval$
which is guaranteed to contain the absolute time of the invocation of $TT.now()$,
i.e. $TT.now().earliest \leq t_{abs}(e) \leq TT.now().latest$, where $e$ denotes
the event of invoking $TT.now()$. More precisely, the absolute time of invocation
of $TT.now()$ is exactly in the middle this $TTinterval$ and spans over
$2\cdot\epsilon$ where $\epsilon$ denotes the instantaneous error bound. $\epsilon$
is derived from applied worst-case local clock drift (assumed to be
$200\mu$s/sec). This instantaneous error bound is adjusted over time for every
single machine.

TrueTime uses GPS and atomic clocks as time references. Google chose these two
types of time reference because the failure modes of them are uncorrelated. A
datacenter typically consists of multiple \emph{time master} machines. Most of
these are GPS receivers with dedicated antennas. All remaining time master
machines use an atomic clock as reference. The references of all time masters are
compared against each other on a regular basis. To avoid incorrect time references
time masters also cross-check their local clock drift against the clock drift of
the reference. If these differ to much, the time master evicts itself.

Each machine has a \emph{timeslave daemon} installed which polls a variety of
time masters, mostly GPS masters but also some Armageddon masters. There is also
a mechanism to detect and reject liars which ensures that timeslave daemons only
synchronize to nonliars.

\begin{table}[ht]
  \centering
  \begin{tabular}{|l||l|}
    \hline
    {\bfseries Method} & {\bfseries Returns} \tabularnewline
    \hline\hline
    $TT.now()$ & $TTinterval$: [$earliest$, $latest$] \tabularnewline
    \hline
    $TT.after(t)$ & true if $t$ has definitely passed \tabularnewline
    \hline
    $TT.before(t)$ & true if $t$ has definitely now arrived \tabularnewline
    \hline
  \end{tabular}
  \caption{Methods provided by the TrueTime API}
  \label{tbl:truetime-api}
\end{table}

\section{Concurrency Control \tbr}
\label{sec:concurrency-control}

The TrueTime API allows Spanner to guarantee some correctness properties of the
implemented concurrency control. This enables the main features in terms of
transaction, i.e. externally-consistent transactions, lock-free read-only
transactions and so-called \emph{snapshot reads}. Snapshot reads are non-blocking
reads in the past at a specific timestamp $t$. Then, Spanner guarantees that
snapshot reads will see the effects of all transactions committed as of $t$.

Spanner provides different operations as can be seen in
Table~\ref{tbl:reads-writes}. Read-only transactions have to be made explicit.
This requires the application/user to predeclare that the read-only transaction
does not include write operations. Moreover, read-only transaction have the
benefits of snapshot isolation because they are \emph{lock-free}. Standalone
writes are simply read-write transaction without any read operations. Non-snapshot
standalone reads are implemented as read-only transactions. Read-only transactions
and snapshot reads proceed on any \emph{sufficiently up-to-date} replica. For
snapshot reads, the application/user has two options. On the one hand, a precise
timestamp can be specified at which the read should be executed. On the other
hand, an upper bound can be specified and Spanner choses a timestamp that
satisfies this bound.

\begin{table}
  \centering
  \begin{tabular}{|l||l|l|}
    \hline
    {\bfseries Operation} & {\bfseries Concurrency Control} &
    {\bfseries Replica required} \tabularnewline
    \hline\hline
    Read-Write transaction & Pessimistic & leader \tabularnewline
    \hline
    Read-Only transaction & Lock-free & leader; any
    \tabularnewline
    \hline
    Snapshot Read, client-provided timestamp & Lock-free & any \tabularnewline
    \hline
    Snapshot Read, client-provided bound & Lock-free & any \tabularnewline
    \hline
  \end{tabular}
  \caption{Reads and writes in Spanner}
  \label{tbl:reads-writes}
\end{table}

\subsection{Paxos Leader Leases \tbr}
\label{subsec:paxos-leader-leases}

The Paxos implementation of Spanner is designed for long-lived leaderships. As
usual in Paxos implementations, a leader is elected by a quorum of lease votes.
If a machine receives such a quorum of lease votes, it gets a lease (10 seconds
by default). The lease vote of a replica is extended automatically on successul
writes.

Spanner enforces a \emph{disjointness invariant} such that for each Paxos group,
each Paxos leader's \emph{lease interval} is disjoint from every other leader's.
In order to guarantee this invariant, Spanner constrains abdication of leaders.
It defines a maximum timestamp used by a leader, denoted as $s_{max}$, which is
advanced by several transactions. A leader must not abdicate before
$TT.after\left(s_{max}\right)$ has passed.

\subsection{Read-Write Transactions \tbr}
\label{subsec:read-write-transactions}

Read-write transaction use 2-phase locking and Spanner may only assign timestamps
in between the two phases. Spanner enforces another invariant, i.e. a
\emph{monotonicity invariant}, which makes use of the disjointness invariant.
This monotonicity invariant guarantees that within each Paxos group, Spanner
assigns timestamps to Paxos writes in monotonically increasing order. To guarantee
this across leaders, Spanner utilizes the disjointness invariant. This implies
that leaders must only assign timestamps within the interval of their respective
leader leases. $s_{max}$ is advanced to $s$ whenever a timestamp $s$ is assigned
by a leader.

To guarantee externally-consistent reads and writes, Spanner enforces a
\emph{external-consistency invariant}. In essence, this invariant guarantees that
whenever the commit of a transaction $T_1$, denoted as $e_1^{commit}$, occurs
before the start of a transaction $T_2$, denoted as $e_2^{start}$, then the commit
timestamp of $T_1$, denoted as $s_1$, is guaranteed to less than the timestamp
of $T_2$, denoted as $s_2$. More formally, the following must hold to guarantee
external consistency:
\begin{center}
$t_{abs}\left(e_1^{commit}\right) < t_{abs}\left(e_2^{start}\right) \Rightarrow s_1 < s_2$.
\end{center}

Spanner obeys two rules in order to guarantee this invariant:
\begin{description}
  \item[Start] Coordinator leader for a write $T_i$ assigns a commit timestamp
    $s_i \geq TT.now().latest$, where $TT.now()$ is executed right after the
    arrival event of commit request for $T_i$.
  \item[Commit Wait] Coordinator leader ensures that clients cannot see data
    committed by $T_i$ until $TT.after\left(s_i\right)$ is true. This ensures
    that the assigned timestamp $s_i$ is less than the absolute commit time of
    $T_i$.~\cite{Corbett:2012}
\end{description}

Figure~\ref{fig:read-write-transaction} illustrates the timeline of a distributed
read-write transaction. There are three participants groups involved, i.e. $P_2$,
$P_3$ and $P_{CL}$ (which is the coordinator leader). In the first step, all three
participants acquire their locks and execute $TT.now()$ right afterwards to obtain
timestamps independently. Every participant logs his timestamp through Paxos and
after this was done, sends its timestamp to the coordinator leader. The
coordinator leader then computes an overall timestamp for the transaction. This
timestamp must be greater than any timestamp the leader has assigned before,
greater or equal to all timestamps received by the participants and greater than
$TT.now().latest$ at the time the coordinator leader received the commit message.
Then, the computed timestamp is logged by the coordinator leader through Paxos
but no participants are notified until the commit wait rule is satisfied. After
commit wait is satisfied, the coordinator leader notifies all other participants
which can release their locks after receiving the notification.

\begin{figure}
  \centering
  \begin{tikzpicture}[thick, scale = 0.5]
    \node (left-end-t1) at (0,2) {};
    \node[right = 9 of left-end-t1] (right-end-t1) {};
    \draw[ultra thick, ->, >=stealth] (left-end-t1.center) -- (right-end-t1.center)
      node[right] (right-end-t1-t) {$t$};
    \node[left = 0 of left-end-t1.west] (left-end-t1-t) {$\mathbf{P_{CL}}$};

    \node[below = 1 of left-end-t1] (left-end-t2) {};
    \node[right = 9 of left-end-t2] (right-end-t2) {};
    \draw[->, >=stealth] (left-end-t2.center) -- (right-end-t2.center)
      node[right] (right-end-t2-t) {$t$};
    \node[left = 0 of left-end-t2.west] (left-end-t2-t) {$P_2$};

    \node[below = 1 of left-end-t2] (left-end-t3) {};
    \node[right = 9 of left-end-t3] (right-end-t3) {};
    \draw[->, >=stealth] (left-end-t3.center) -- (right-end-t3.center)
      node[right] (right-end-t3-t) {$t$};
    \node[left = 0 of left-end-t3.west] (left-end-t3-t) {$P_3$};

    \node[right = 0.25 of left-end-t1] (acquire-locks-t1) {};
    \node[right = 1.05 of left-end-t2] (acquire-locks-t2) {};
    \node[right = 0.65 of left-end-t3] (acquire-locks-t3) {};
    \foreach \x in {1, 2, 3}{
      \draw[blue] (acquire-locks-t\x.center) -- ++(0, -0.25);
      \draw[blue] (acquire-locks-t\x.center) -- ++(0, 0.25);
    }

    \node[above = 0.75 of acquire-locks-t1]
      (acquire-locks-t) {\scriptsize\color{blue}Acquire all locks};

    \foreach \x in {1, 2, 3}{
      \node[right = -0.1 of acquire-locks-t\x] (compute-s-t\x) {};
    }
    \node[below = 1 of acquire-locks-t3.center]
      (compute-s-t) {\scriptsize $s=TT.now()$};
    \draw[<-, >=stealth, dotted] (compute-s-t1.center) -- ++(0, -7);
    \draw[<-, >=stealth, dotted] (compute-s-t2.center) -- ++(0, -4.6);
    \draw[<-, >=stealth, dotted] (compute-s-t3.center) -- ++(0, -2);

    \node[right = 1 of compute-s-t2] (start-log-t2) {};
    \node[right = 1.75 of compute-s-t3] (start-log-t3) {};
    \node[above = 2 of start-log-t2] (start-log-t)
      {\scriptsize Start Logging};
    \draw[<-, >=stealth, dotted] (start-log-t2.center) -- ++(0, 4.25);
    \draw[<-, >=stealth, dotted] (start-log-t3.center) -- ++(0, 6.75);

    \node[right = 1 of start-log-t3] (end-log-t3) {};
    \node[right = 1.75 of start-log-t2] (end-log-t2) {};
    \node[above = 2 of end-log-t2] (end-log-t)
      {\scriptsize Logging done};
    \draw[<-, >=stealth, dotted] (end-log-t2.center) -- ++(0, 4.25);
    \draw[<-, >=stealth, dotted] (end-log-t3.center) -- ++(0, 6.75);
    \node[right = 5.5 of compute-s-t1] (compute-s-global) {};
    \draw[<-, >=stealth, black!50!green] (compute-s-global) ++(-0.1, 0) --
      (end-log-t3.north east);
    \draw[<-, >=stealth, black!50!green] (compute-s-global) ++(-0.65, 0) --
      (end-log-t2.north east);
    \draw[black!50!green, ultra thick] (compute-s-global.center) -- ++(0, -0.25);
    \draw[black!50!green, ultra thick] (compute-s-global.center) -- ++(0, 0.25);

    \node[above = 0.25 of compute-s-global] (compute-s-global-t)
      {\scriptsize\color{black!50!green} Overall $s$};
    \node[below = 0 of end-log-t3] (prepared-t)
      {\scriptsize\color{black!50!green}Prepared; Send $s$};

    \node[right = 0.5 of compute-s-global] (commit-wait-done) {};
    \node[below = 3.45 of commit-wait-done] (commit-wait-done-t)
      {\scriptsize Commit Wait done};
    \draw[<-, >=stealth, dotted] (commit-wait-done.center) -- ++(0, -7);
    \node[right = 0 of commit-wait-done] (release-locks-t1) {};
    \draw[red] (release-locks-t1.center) -- ++(0, -0.25);
    \draw[red] (release-locks-t1.center) -- ++(0, 0.25);
    \node[above right = 0.75 and -0.75 of release-locks-t1]
      (release-locks-t1-t) {\scriptsize\color{red}Release all locks};

    \node[right = -0.15 of release-locks-t1] (notify-t1) {};
    \node[right = 3.1 of end-log-t2] (notify-t2) {};
    \node[right = 3.35 of end-log-t3] (notify-t3) {};
    \foreach \x in {2, 3}{
      \draw[->, >=stealth, black!50!green] (notify-t1.center) --
        (notify-t\x.center); 
    }

    \node[right = -0.15 of notify-t2] (release-locks-t2) {};
    \node[right = -0.15 of notify-t3] (release-locks-t3) {};
    \draw[red] (release-locks-t2.center) -- ++(0, -0.25);
    \draw[red] (release-locks-t2.center) -- ++(0, 0.25);
    \draw[red] (release-locks-t3.center) -- ++(0, -0.25);
    \draw[red] (release-locks-t3.center) -- ++(0, 0.25);
  \end{tikzpicture}
  \caption{Read-write transaction}
  \label{fig:read-write-transaction}
\end{figure}

This also explains why, according to Table~\ref{tbl:reads-writes}, Read-write
transactions only involve the \emph{leader} replicas. This is because only the
leader replicas of each participant communicate with the coordinator leader.
Hence, no other replicas are involved.

\subsection{Snapshot Reads \tbr}
\label{subsec:reads-at-timestamp}

In order to provide snapshot reads, we first need to define
\emph{sufficiently up-to-date}. Therefore Spanner tracks a \emph{safe time},
denoted as $t_{safe}$, at each replica. $t_{safe}$ is defined as the maximum
timestamp at which the corresponding replica is up-to-date. Hence, a replica
satisfies reads at every timestamp $t$ if $t \leq t_{safe}$ holds. Spanner can
then perform a snapshot read at any replica that is sufficiently up-to-date. As
a consequence, snapshot reads have the same performance benefits as provided by
snapshot isolation since they are lock-free.

The safe time is determined by the safe time of a Paxos state machine, denoted
as $t_{safe}^{Paxos}$, and the safe t ime of the transaction manager, i.e.
$t_{safe}^{TM}$. The overall safe time of a replica then is the minimum of
these two. $t_{safe}^{Paxos}$ is simple equal to the time of the highest-applied
Paxos write because if the Paxos level is reached in the spanserver software stack,
then the corresponding transaction has already been committed. $t_{safe}^{TM}$
is $\infty$ if there exist distributed transaction which are in between the two
phases of two phases of the 2-phase commit because in this case the system does
not guaranteed that this transaction will commit.

\subsection{Read-Only Transactions \tbr}
\label{subsec:read-only-transactions}

Read-only transactions are also lock-free and execute in two phases. In the first
phase, a timestamp $s_{read}$ is assigned. In the second phase, the read-only
transaction is executed as a snapshot read at $s_{read}$. Spanner assigns
the oldest timestamp which preserves external consistency to $s_{read}$.

In general, read-only transactions involve group leaders and they may also involve
any other replica which is sufficiently up-to-date. Group leaders are involved
in the negotiation phase. In this phase, all involved Paxos groups negotiate in
order to determine $s_{read}$. To accomplish this, Spanner requires every
read-only transaction to define a \emph{scope} expression which summarizes all
keys that will be read by the respective transaction.

If all values of the scope can be served by a single Paxos group, the client
issues the read-only transaction to the corresponding group leader. The group
leader then assign the timestamp of the last committed write to $s_{read}$ and
executes the transaction.

If serving the values of the scope involves multiple Paxos groups, Spanner avoids
a negotiation round (which would be one viable options). Instead, Spanner simply
assigns$TT.now().latest$ to $s_{read}$, even though this may have to wait for
$t_{safe}$ to advance.

In both cases, the snapshot read executed at $s_{read}$ can execute on any
replica. The leader replicas are only involved in the negotiation phase.

\subsection{Schema-Change Transactions \tbr}
\label{subsec:schema-change-transactions}

Schema-change transactions are variants of standard transactions which change
schemas atomically without blocking the whole system.

There are two phases involved. In the first phase, the \emph{prepare} phase,
Spanner explicitly assigns a future timestamp to the schema-change transaction.
In the phase, reads and writes synchronize with any registered schema-change
timestamp at time $t$. Of course, this concerns only the reads and writes which
implicitly depend on this schema. Any read/write transaction may proceed if
its timestamp preceeds $t$. If its timestamp is after $t$, the corresponding
read/write has to block behind the schema-change transaction. This is essentially
enabled by the TrueTime API because without TrueTime, it would not be guaranteed
that the timestamp of the schema-change transaction is in the future.

\section{Performance \tbr}
\label{sec:performance}

\begin{table}[ht]
  \centering
  { \footnotesize
    \begin{tabular}{|l||l|l|l||l|l|l|}
      \hline
      & \multicolumn{3}{c||}{{\bfseries Latency in [ms] (less is better)}} &
      \multicolumn{3}{c|}{{\bfseries Throughput in [Kops/sec] (more is better)}}
      \tabularnewline
      \cline{2-7}
      {\bfseries Replicas} & {\bfseries Write} & {\bfseries RO Transaction} &
      {\bfseries Snapshot Read} & {\bfseries Write} &
      {\bfseries RO Transaction} & {\bfseries Snapshot Read} \tabularnewline
      \hline\hline
      1D & $9.4\pm0.6$ & - & - & $4.0\pm0.3$ & - & - \tabularnewline
      \hline
      1 & $14.4\pm1.0$ & $1.4\pm0.1$ & $1.3\pm0.1$ & $4.1\pm0.05$ &
      $10.9\pm0.4$ & $13.5\pm0.1$ \tabularnewline
      \hline
      3 & $13.9\pm0.6$ & $1.3\pm0.1$ & $1.2\pm0.1$ & $2.2\pm0.5$ &
      $13.8\pm3.2$ & $38.5\pm0.3$ \tabularnewline
      \hline
      5 & $14.4\pm0.4$ & $1.4\pm0.05$ & $1.3\pm0.04$ & $2.8\pm0.3$ &
      $25.3\pm5.2$ & $50.0\pm1.1$ \tabularnewline
      \hline
    \end{tabular}
  }
  \caption{Performance of operations. Mean and standard deviation over 10
    runs.~\cite{Corbett:2012}
  }
  \label{tbl:performance-of-operations}
\end{table}

This section summarizes the core results of the evaluation of Spanner, i.e. the
latency and throughput performance as well as some results about availability
and the TrueTime API.

Table~\ref{tbl:performance-of-operations} summarizes the latency and throughput
performance of Spanner. 1D means one replica with commit wait disabled. Hence,
only write transactions are evaluated because commit wait is only involved in such
transactions. As we can see, the latency of all transaction types is almost stable
for an increasing number of replicas whereas the standard deviation decreases.
The throughput of read-only transaction increases with the number of replicas
because the number of spanservers increases (in the experiments, the number of
replicas and the number of spanservers were the same. The throughput of write
transactions also benefits from this fact. However, the number of write operations
to perform increases linearly with the number of replicas. Snapshot reads benefit
the most from the Spanner architecture. The throuhgput of snapshot reads increases
almost linear with the number of replicas.

The availability of Spanner is only affected if zones containing multiple leaders
are shut down without notifying the leader upfront. In this case, it takes Spanner
about 10s to recover full availability. If the leaders are notified to abdicate
before their zone is shut down, the availability stays almost unaffected. The same
if true if a zone without a leader is shut down.

Google also provided results for some experiments concerning TrueTime. All
assumptions Google made when designing Spanner would break if the local clock
drift would exceed $200\mu$s/sec. But Google found that clock issues are extremely
infrequent relative to more serious hardware problems, e.g. bad CPUs. Moreover,
Google provided empirical results on the values of $\epsilon$. Typically,
$\epsilon$ is never larger than 10ms. It decreases as the network infrastructure
is improved and increases if time masters go offline.

\section{Future Work \tbr}
\label{sec:future-work}

Google plans to implement optimistic parallel reads but it turns out that
implementing this is non-trivial. They also want to support direct configuration
changes of the Paxos state machines and to reduce $\epsilon$ to be $\leq 1ms$ to
improve the performance if data is sharded between nearby datacenters.

Moreover, the node-local performance can be improved for complex queries. By now,
node-local data structures are designed for key-value accesses and these could be
improved by algorithms and data structure from database literature.

To improve the performance even further, Google plans to move processes between
datacenters if the data they operate on is moved. This would have a high impact
because the performance of the processes is close to optimal if the data is as
local as possible. However, moving processes between datacenters raises even more
complex problems.

\section{Discussion \tbr}
\label{sec:discussion}

The system introduced by Google is definitly impressive because they were able
to establish a notion of global wall clock time (that's how they call it in their
presentation) which enabled them to guarantee serialization order for
globally-distributed transactions. To the best of the author's knowledge, Spanner
was the first system to provide much stronger consistency and time semantics while
data is sharded and replicated at global scale. Google also proved all
'\emph{Relational databases are dead}' people wrong even though systems like
Bigtable helped creating the NoSQL trend. Reading about a system like Spanner is
definitly more than a vital sign of relational database systems. From a
developer's point of view, this is definitly interesting because Spanner provides
the more intuitive schematized view of tables. In contrast to NOSQL database
systems, Spanner could be used as replacement for other relational database
systems, e.g. MySQL, directly. Spanner is also highly flexible to schema changes
because of the atomic schema change transactions.

However, I believe that rebuilding a Spanner-like system is difficult for
small- to mid-sized companies because there is a lot of hardware involved, for
example all the GPS and atomic clocks. Unless this hardware infrastructure is
not available it is also hard to reimplement the TrueTime API. Nonetheless,
companies which need a globally-distributed database with all the guarantees
Spanner provides are likely to be almost as big as Google. Hence, such companies
may also be able to build a Spanner-like system. If every datacenter would provide
GPS and Armageddon time masters independently, the reimplementation of such a
system would be much easier.

Another part that may influence the performance of Spanner is the distributed
file system, i.e. Colossus, the successor of the Google File System (GFS). I did
not find reliable information about the implementation or the performance of
Colossus. It is unknown how much impact this distributed file system has on the
performance of Spanner. Thus, it is hard to say if a reimplementation of Spanner
would be able to compete with Spanner's performance.

Of course, such a system may also be built upon e.g. the Network Time Protocol
(NTP). But this would result in poor performance compared to Spanner because the
error interval would be much larger compared to $2\cdot\epsilon$. And besides
providing distributed general-purpose transactions, the performance of Spanner
makes the whole system that impressive. A system using e.g. NTP would never be
able to achieve this performance while guaranteeing serialization order. Spanner
can synchronize multiple systems without any communication between these systems.
NTP-based systems would have to communicate with each other. For systems far apart
such a system would never be able to compete with Spanner.

Spanner is robust to GPS blackouts because of the usage ofatomic clocks, even
though the performance of Spanner may suffer due to high values of $\epsilon$.
Would be interesting to see how such a blackout affects Spanner's performance.

It would also be interesting how failures of the two singletons per universe,
i.e. universemaster and placement driver, affect Spanner. Since the universemaster
is just a debugging console, most probably a failure in this component will not
affect Spanner at all. This would probably be a totally different story if the
placement driver does not operate correctly. First of all, there would be no
load balancing between zones. Hence, zones could be run out of space if the
placement driver is down long enough. Furthermore, updated replication
constraints, would not be propagated since the placement driver is responsible
to move the data if the replication constraints are modified.

In terms of applications, applications which read a lot but write rarely or little
data would benefit the most from a system like Spanner. If an application uses
lots of replicas which are spread across the globe and writes large amounts of data
regularly, Spanner may not provide the performance mentioned in the original
paper. However, Spanner may still outperform both, Megastore and Bigtable while
providing much stronger consistency guarantees.

%\clearpage
\bibliographystyle{abbrv}
\bibliography{paper}

\end{document}

### 2016-02-16 paper/paper.tex
Iterate Introduction

@@
 
-\section{Introduction \tbr}
+\section{Introduction}
 \label{sec:introduction}
@@
 globally-distributed database referred to as \emph{Spanner}. This system is based
-upon Paxos~\cite{Lamport:1998} state machines on machines spread across multiple
-datacenters across the world. This global replication guarantees high availability
-and also implies some desirable locality properties which benefit the performance.
+upon Paxos~\cite{Lamport:1998} state machines running on machines spread across
+multiple datacenters across the world. This global replication guarantees high
+availability and also implies some desirable locality properties which benefit
+the performance.
 
@@
 distributed database. According to Google, it is difficult for applications/users
-to deal with complex and evolving schemas. Moreover, Bigtable only guarantees to
-be eventually-consistent. In contrast, Spanner provides much stronger consistency
-guarantees. On the other hand, Megastore suffers from poor write throughput even
-though it is used for many applications.
+to deal with complex and evolving schemas in Bigtable. Moreover, Bigtable only
+guarantees to be eventually-consistent. In contrast, Spanner provides much
+stronger consistency guarantees. On the other hand, Megastore suffers from poor
+write throughput even though many applications use it.
 
-Therefore, Google decided to introduce another distributed, highly available
-database system which evolved from a key-value store somehow similar to the one
-used by Bigtable (but we will see that this is only partially correct). A core
-feature provided by Spanner are so-called \emph{externally-consistent} distributed
-general-purpose transactions. Of course, Spanner also provides an SQL-like query
-language to enable applications/users to interact with Spanner in a well-known
-way. The data is stored in schematized \emph{semi-relational} tables and the data
-is versioned automatically. Google also implemented a garbage collection to deal
-with old data. The replication can be dynamically configured in order to control
-the locality of the replicas. This is useful to control write and read latency
-as well as availability. Another main contribution of Spanner are
-\emph{globally-consistent} reads at a specific timestamp. This and
-externally-consistent reads and writes enable Spanner to provide a special type
-of transaction, so-called \emph{schema-change} transactions, which provides atomic
-schema changes without locking the whole database and all ongoing transactions.
+Therefore, Google decided to introduce another distributed, highly available and
+scalable database system which evolved from a key-value store somehow similar to
+the one used in Bigtable (but we will see that this is only partially correct).
+A core feature provided by Spanner are so-called \emph{externally-consistent}
+distributed general-purpose transactions. Of course, Spanner also provides an
+SQL-like query language to enable applications/users to interact with Spanner in
+a familiar way. The data is stored in schematized \emph{semi-relational} tables
+and the data is versioned automatically. Google also implemented a garbage
+collection to deal with old data. The replication can be dynamically configured
+in order to control the locality of replicas. This is useful to control write
+and read latency as well as availability. Another main contribution of Spanner
+are high-performance \emph{globally-consistent} reads at a specific timestamp.
+This and externally-consistent reads and writes enable Spanner to provide a
+special type of transaction, so-called \emph{schema-change} transactions, which
+provide atomic schema changes without blocking the whole database.
 

### 2016-02-16 paper/paper.tex
Iterate Implementation introduction

@@
 
-\section{Implementation \tbr}
+\section{Implementation}
 \label{sec:implementation}
@@
 multiple zones. A universe has two global components: A \emph{universemaster}
-which, essentially, is a debugging console and shows the status of each zone. A
+which is a debugging console and shows the status of each zone. A
 \emph{placement driver} which automatically moves data between zones due to
@@
 One can also think of a zone as geographically wide-spread locations, e.g.
-continents. Using such a physical isolation, Spanner is resistent to events which
-affect large areas, e.g. natural disasters or war. Hence, a well-configured
+continents. Using this kind of physical isolation, Spanner is resistent to events
+which affect large areas, e.g. natural disasters or war. Hence, a well-configured
 Spanner deployment is available almost all the time.

### 2016-02-16 paper/paper.tex
Iterate Spanserver Software Stack

@@
 
-Each spanserver also consists of multiple software layers (four at least). At the
-bottom layer, every spanserver is responsible for between 100 and 1,000
+Each spanserver again consists of multiple software layers (four at least). At
+the bottom layer, every spanserver is responsible for between 100 and 1,000
 \emph{tablets}. A tablet is bag of key-value mappings of the form
@@
 On top of each tablet, a Paxos state machine is implemented in order to support
-consistent replication. One Paxos state machine is elect as leader. Leases are
-assigned in a long-lived manner (default lease length is 10 seconds). A leader
+consistent replication. One Paxos state machine is elect as leader. Leader leases
+are assigned in a long-lived manner (default lease length is 10 seconds). A leader
 implements three additional layers. A \emph{lock table} which stores the state
@@
 referred to as \emph{coordinator leader}.
-Figure~\ref{subfig:spanserver-software-stack} shows the spanserver software stack
-as described before.
+Figure~\ref{subfig:spanserver-software-stack} shows the spanserver software stack.
 

### 2016-02-16 paper/paper.tex
Iterate Directories

@@
 illustrates this principle. Directory 2 is moved from Paxos group 1 to Paxos group
-2 and at the same time from zone 2 to zone 3. Google shortly mentions that this
+2 and, at the same time, from zone 2 to zone 3. Google shortly mentions that this
 is an over-simplified view because a directory is split into multiple
-\emph{fragments} if it grows too large. However, Google does not provide more
-details.
+\emph{fragments} if it grows too large. However, Google does not provide any
+additional details.
 
@@
 be described later, the concept of directories enables Google to improve the
-performance of Spanner by placing directories in the same Paxos group which are
-frequently accessed together. The fact that in general multiple directories are
-placed in the same Paxos group is another difference between the key-value mapping
-of Spanner and the one used in Bigtable. A Spanner tablet may contain multiple
-contiguous ranges of keys (= directories) as opposed to a Bigtable tablet which
-may only contain one contiguous range of keys. Figure~\ref{fig:comparison-tablets}
-shows an example comparison of the two tablet types.
+performance of Spanner by placing directories which are frequently accessed
+together in the same Paxos group. The fact that in general multiple directories
+are placed in the same Paxos group is another difference between the key-value
+mapping of Spanner and the one used in Bigtable. A Spanner tablet may contain
+multiple contiguous ranges of keys (= directories) as opposed to a Bigtable tablet
+which may only contain one contiguous range of keys.
+Figure~\ref{fig:comparison-tablets} shows an example comparison of the two tablet
+types.
 
@@
   \end{subfigure}
-  \caption{Comparison of Spanner's and Bigtable's tablets}
+  \caption{Comparison of Spanner's and Bigtable's tablet}
   \label{fig:comparison-tablets}
@@
 application/user can specify geographic-replication properties. An application
-can control replication in two dimension, i.e. the number and types of replicas,
-and, the geographic placement of replicas.
+can control replication in two dimensions, namely the number and types of
+replicas, and the geographic placement of replicas.
 

### 2016-02-16 paper/paper.tex
Iterate Directories
Iterate Data Model

@@
 
-\subsection{Directories \tbr}
+\subsection{Directories}
 \label{subsec:directories}
@@
 
-\subsection{Data Model \tbr}
+\subsection{Data Model}
 \label{subsec:data-model}
@@
 all tables of a hierarchy which belong to the same key of the topmost table of
-the hierarchy. Figure~\ref{subfig:interleaved-tables} shows an examples of this
+the hierarchy. Figure~\ref{subfig:interleaved-tables} shows an example of this
 concept. In this example, the \texttt{Users} table is the topmost table of the
 hierarchy and the \texttt{Albums} table is one level below. As we can see, all
-rows of \texttt{Albums} starting with \texttt{1} are located in the same directory
-(\texttt{3365}). This directory has a common prefix, i.e. \texttt{1}, from the
-\texttt{Users} table. The same principle can be seen for key \texttt{2} of the
-\texttt{Users} table. This interleaving of tables exploits locality relationships
-of multiple tables and therefore is significant for Spanner's performance.
+rows of \texttt{Albums} connected to user \texttt{1} are located in the same
+directory (\texttt{3365}). This directory has a common prefix, i.e. \texttt{1},
+from the \texttt{Users} table. The same principle can be seen for key \texttt{2}
+of the \texttt{Users} table. This interleaving of tables exploits locality
+relationships of multiple tables and therefore is significant for Spanner's
+performance.
 

### 2016-02-16 paper/paper.tex
Iterate TrueTime

@@
 
-\section{TrueTime \tbr}
+\section{TrueTime}
 \label{sec:truetime}
@@
 This section describes the key enabler of Spanner, i.e. the novel \emph{TrueTime}
-API. Essentially, this API enables Google to provide timestamps for transactions
-which reflect serialization order.
+API. Essentially, this API enables Google to assign timestamps to distributed
+transactions which reflect serialization order.
 
@@
 the event of invoking $TT.now()$. More precisely, the absolute time of invocation
-of $TT.now()$ is exactly in the middle this $TTinterval$ and spans over
-$2\cdot\epsilon$ where $\epsilon$ denotes the instantaneous error bound. $\epsilon$
-is derived from applied worst-case local clock drift (assumed to be
-$200\mu$s/sec). This instantaneous error bound is adjusted over time for every
-single machine.
+of $TT.now()$ is exactly in the middle this $TTinterval$. The interval itself
+spans over $2\cdot\epsilon$, where $\epsilon$ denotes the instantaneous error
+bound. $\epsilon$ is derived from applied worst-case local clock drift (assumed
+to be $200\mu$s/sec). This instantaneous error bound is adjusted over time for
+every single machine.
 
 TrueTime uses GPS and atomic clocks as time references. Google chose these two
-types of time reference because the failure modes of them are uncorrelated. A
+types of time references because the failure modes of them are uncorrelated. A
 datacenter typically consists of multiple \emph{time master} machines. Most of
 these are GPS receivers with dedicated antennas. All remaining time master
-machines use an atomic clock as reference. The references of all time masters are
-compared against each other on a regular basis. To avoid incorrect time references
-time masters also cross-check their local clock drift against the clock drift of
-the reference. If these differ to much, the time master evicts itself.
+machines use an atomic clock as reference. Machines using an atomic clock as time
+reference are referred to as \emph{Armageddon} masters. The references of all time
+masters are compared against each other on a regular basis. To avoid incorrect
+time references, time masters also cross-check their local clock drift against
+the clock drift of the reference. If these differ too much, the time master evicts
+itself.
 
@@
     \hline
-    $TT.before(t)$ & true if $t$ has definitely now arrived \tabularnewline
+    $TT.before(t)$ & true if $t$ has definitely not arrived \tabularnewline
     \hline
   \end{tabular}
-  \caption{Methods provided by the TrueTime API}
+  \caption{Methods provided by the TrueTime API.}
   \label{tbl:truetime-api}
@@
 
-The system introduced by Google is definitly impressive because they were able
+The system introduced by Google is definitely impressive because they were able
 to establish a notion of global wall clock time (that's how they call it in their
@@
 Bigtable helped creating the NoSQL trend. Reading about a system like Spanner is
-definitly more than a vital sign of relational database systems. From a
-developer's point of view, this is definitly interesting because Spanner provides
+definitely more than a vital sign of relational database systems. From a
+developer's point of view, this is interesting because Spanner provides
 the more intuitive schematized view of tables. In contrast to NOSQL database

### 2016-02-16 paper/paper.tex
Iterate Concurrency Control introduction

@@
 
-\section{Concurrency Control \tbr}
+\section{Concurrency Control}
 \label{sec:concurrency-control}
@@
 The TrueTime API allows Spanner to guarantee some correctness properties of the
-implemented concurrency control. This enables the main features in terms of
-transaction, i.e. externally-consistent transactions, lock-free read-only
-transactions and so-called \emph{snapshot reads}. Snapshot reads are non-blocking
-reads in the past at a specific timestamp $t$. Then, Spanner guarantees that
-snapshot reads will see the effects of all transactions committed as of $t$.
+implemented concurrency control. These enable some of the main features, i.e.
+externally-consistent transactions, lock-free read-only transactions and so-called
+\emph{snapshot reads}. Snapshot reads are non-blocking reads in the past at a
+specific timestamp $t$. In this case, Spanner guarantees that snapshot reads will
+see the effects of all transactions committed as of $t$.
 
-Spanner provides different operations as can be seen in
+Spanner provides different operations, as can be seen in
 Table~\ref{tbl:reads-writes}. Read-only transactions have to be made explicit.
@@
   \end{tabular}
-  \caption{Reads and writes in Spanner}
+  \caption{Reads and writes in Spanner.}
   \label{tbl:reads-writes}

### 2016-02-16 paper/paper.tex
Iterate Paxos Leader Lease
Iterate Read-Write Transactions

@@
 
-\subsection{Paxos Leader Leases \tbr}
+\subsection{Paxos Leader Leases}
 \label{subsec:paxos-leader-leases}
@@
 If a machine receives such a quorum of lease votes, it gets a lease (10 seconds
-by default). The lease vote of a replica is extended automatically on successul
+by default). The lease vote of a replica is extended automatically on successful
 writes.
@@
 
-\subsection{Read-Write Transactions \tbr}
+\subsection{Read-Write Transactions}
 \label{subsec:read-write-transactions}
 
-Read-write transaction use 2-phase locking and Spanner may only assign timestamps
-in between the two phases. Spanner enforces another invariant, i.e. a
+Read-write transactions use two-phase locking and Spanner may only assign
+timestamps in between the two phases. Spanner enforces another invariant, i.e. a
 \emph{monotonicity invariant}, which makes use of the disjointness invariant.
@@
 this across leaders, Spanner utilizes the disjointness invariant. This implies
-that leaders must only assign timestamps within the interval of their respective
-leader leases. $s_{max}$ is advanced to $s$ whenever a timestamp $s$ is assigned
-by a leader.
+that a leader must only assign timestamps within the interval of its leader
+lease. $s_{max}$ is advanced to $s$ whenever a timestamp $s$ is assigned by a
+leader.
 
-To guarantee externally-consistent reads and writes, Spanner enforces a
-\emph{external-consistency invariant}. In essence, this invariant guarantees that
+To guarantee externally-consistent reads and writes, Spanner enforces an
+\emph{external consistency invariant}. In essence, this invariant guarantees that
 whenever the commit of a transaction $T_1$, denoted as $e_1^{commit}$, occurs
 before the start of a transaction $T_2$, denoted as $e_2^{start}$, then the commit
-timestamp of $T_1$, denoted as $s_1$, is guaranteed to less than the timestamp
-of $T_2$, denoted as $s_2$. More formally, the following must hold to guarantee
-external consistency:
+timestamp of $T_1$, denoted as $s_1$, is guaranteed to be smaller than the
+timestamp of $T_2$, denoted as $s_2$. More formally, the following must hold to
+guarantee external consistency:
 \begin{center}
-$t_{abs}\left(e_1^{commit}\right) < t_{abs}\left(e_2^{start}\right) \Rightarrow s_1 < s_2$.
+$t_{abs}\left(e_1^{commit}\right) < t_{abs}\left(e_2^{start}\right) \Rightarrow s_1 < s_2$
+\qquad\qquad $t_{abs}(e)$ \ldots absolute time of an event $e$
 \end{center}
 
+\noindent
 Spanner obeys two rules in order to guarantee this invariant:
@@
     $s_i \geq TT.now().latest$, where $TT.now()$ is executed right after the
-    arrival event of commit request for $T_i$.
+    arrival event of the commit request for $T_i$.
   \item[Commit Wait] Coordinator leader ensures that clients cannot see data
@@
 timestamp must be greater than any timestamp the leader has assigned before,
-greater or equal to all timestamps received by the participants and greater than
+greater or equal to all timestamps received from the participants and greater than
 $TT.now().latest$ at the time the coordinator leader received the commit message.
@@
 commit wait is satisfied, the coordinator leader notifies all other participants
-which can release their locks after receiving the notification.
+which are allowed to release their locks after receiving this notification.
 
@@
   \end{tikzpicture}
-  \caption{Read-write transaction}
+  \caption{Timeline of a read-write transaction.}
   \label{fig:read-write-transaction}
@@
 This also explains why, according to Table~\ref{tbl:reads-writes}, Read-write
-transactions only involve the \emph{leader} replicas. This is because only the
-leader replicas of each participant communicate with the coordinator leader.
-Hence, no other replicas are involved.
+transactions only involve the \emph{leader} replicas. Only the leader replicas
+of each participant communicate with the coordinator leader. Hence, no other
+replicas are involved.
 

### 2016-02-16 paper/paper.tex
Iterate Snapshot Reads

@@
 
-\subsection{Snapshot Reads \tbr}
+\subsection{Snapshot Reads}
 \label{subsec:reads-at-timestamp}
@@
 a consequence, snapshot reads have the same performance benefits as provided by
-snapshot isolation since they are lock-free.
+snapshot isolation since they are totally lock-free.
 
 The safe time is determined by the safe time of a Paxos state machine, denoted
-as $t_{safe}^{Paxos}$, and the safe t ime of the transaction manager, i.e.
+as $t_{safe}^{Paxos}$, and the safe time of the transaction manager, i.e.
 $t_{safe}^{TM}$. The overall safe time of a replica then is the minimum of
-these two. $t_{safe}^{Paxos}$ is simple equal to the time of the highest-applied
-Paxos write because if the Paxos level is reached in the spanserver software stack,
-then the corresponding transaction has already been committed. $t_{safe}^{TM}$
-is $\infty$ if there exist distributed transaction which are in between the two
-phases of two phases of the 2-phase commit because in this case the system does
-not guaranteed that this transaction will commit.
+these two.
+
+$t_{safe}^{Paxos}$ is simply assigned the time of the highest-applied Paxos write
+because if the Paxos level is reached in the spanserver software stack, then the
+corresponding transaction has already been committed. $t_{safe}^{TM}$ is $\infty$
+if there exists a distributed transaction which is in between the two phases of
+the two-phase commit because in this case the system does not know if this
+transaction will commit.
 

### 2016-02-16 paper/paper.tex
Iterate Read-Only Transactions

@@
 
-\subsection{Read-Only Transactions \tbr}
+\subsection{Read-Only Transactions}
 \label{subsec:read-only-transactions}
@@
 
-In general, read-only transactions involve group leaders and they may also involve
-any other replica which is sufficiently up-to-date. Group leaders are involved
-in the negotiation phase. In this phase, all involved Paxos groups negotiate in
-order to determine $s_{read}$. To accomplish this, Spanner requires every
-read-only transaction to define a \emph{scope} expression which summarizes all
-keys that will be read by the respective transaction.
+In general, read-only transactions involve group leaders but they may also involve
+any other sufficiently up-to-date replica in the process. Group leaders are
+involved in the negotiation phase. In this phase, all involved Paxos groups
+negotiate in order to determine $s_{read}$. To accomplish this, Spanner requires
+every read-only transaction to define a \emph{scope} expression which summarizes
+all keys that will be read by the respective transaction.
 
@@
 issues the read-only transaction to the corresponding group leader. The group
-leader then assign the timestamp of the last committed write to $s_{read}$ and
+leader then assigns the timestamp of the last committed write to $s_{read}$ and
 executes the transaction.
@@
 If serving the values of the scope involves multiple Paxos groups, Spanner avoids
-a negotiation round (which would be one viable options). Instead, Spanner simply
-assigns$TT.now().latest$ to $s_{read}$, even though this may have to wait for
+a negotiation round (which would be a viable option). Instead, Spanner simply
+assigns $TT.now().latest$ to $s_{read}$, even though this may have to wait for
 $t_{safe}$ to advance.
 
-In both cases, the snapshot read executed at $s_{read}$ can execute on any
-replica. The leader replicas are only involved in the negotiation phase.
+In both cases, the snapshot read at $s_{read}$ can execute on any sufficiently
+up-to-date replica. The leader replicas are only involved in the negotiation
+phase.
 

### 2016-02-16 paper/paper.tex
Iterate Schema-Change Transactions

@@
 
-\subsection{Schema-Change Transactions \tbr}
+\subsection{Schema-Change Transactions}
 \label{subsec:schema-change-transactions}
@@
 Schema-change transactions are variants of standard transactions which change
-schemas atomically without blocking the whole system.
+schemas atomically without blocking any other transactions.
 
@@
 Spanner explicitly assigns a future timestamp to the schema-change transaction.
-In the phase, reads and writes synchronize with any registered schema-change
-timestamp at time $t$. Of course, this concerns only the reads and writes which
-implicitly depend on this schema. Any read/write transaction may proceed if
-its timestamp preceeds $t$. If its timestamp is after $t$, the corresponding
-read/write has to block behind the schema-change transaction. This is essentially
-enabled by the TrueTime API because without TrueTime, it would not be guaranteed
-that the timestamp of the schema-change transaction is in the future.
+In the subsequent phase, reads and writes synchronize with any registered
+schema-change timestamp at time $t$. Of course, this affects only reads and
+writes which implicitly depend on the modified schema. Any read/write transaction
+may proceed if its timestamp preceeds $t$. If its timestamp is after $t$, the
+corresponding read/write has to block behind the schema-change transaction. This
+is essentially enabled by the TrueTime API because without TrueTime, it would
+not be guaranteed that the timestamp of the schema-change transaction is in the
+future on every single machine.
 

### 2016-02-16 paper/paper.tex
Iterate Performance

@@
 
-\section{Performance \tbr}
+\section{Performance}
 \label{sec:performance}
@@
 This section summarizes the core results of the evaluation of Spanner, i.e. the
-latency and throughput performance as well as some results about availability
+latency and throughput performance as well as some results concerning availability
 and the TrueTime API.
@@
 Table~\ref{tbl:performance-of-operations} summarizes the latency and throughput
-performance of Spanner. 1D means one replica with commit wait disabled. Hence,
-only write transactions are evaluated because commit wait is only involved in such
-transactions. As we can see, the latency of all transaction types is almost stable
-for an increasing number of replicas whereas the standard deviation decreases.
-The throughput of read-only transaction increases with the number of replicas
-because the number of spanservers increases (in the experiments, the number of
-replicas and the number of spanservers were the same. The throughput of write
+performance of Spanner. 1D means one replica with Commit Wait disabled. Hence,
+only write transactions are evaluated because Commit Wait is not involved in any
+other type of transaction. As we can see, the latency of all transaction types is
+almost stable for an increasing number of replicas whereas the standard deviation
+decreases. The throughput of read-only transactions increases with the number of
+replicas because the number of spanservers increases. In the experiments, the
+number of replicas and the number of spanservers were equal. The throughput of write
 transactions also benefits from this fact. However, the number of write operations
-to perform increases linearly with the number of replicas. Snapshot reads benefit
-the most from the Spanner architecture. The throuhgput of snapshot reads increases
-almost linear with the number of replicas.
+to perform increases linearly with the number of replicas. This explains why the
+throughput of write transactions does not increase with the number of replicas.
+Snapshot reads benefit the most from the Spanner architecture. The throughput of
+snapshot reads increases almost linearly with the number of replicas.
 
 The availability of Spanner is only affected if zones containing multiple leaders
-are shut down without notifying the leader upfront. In this case, it takes Spanner
-about 10s to recover full availability. If the leaders are notified to abdicate
-before their zone is shut down, the availability stays almost unaffected. The same
-if true if a zone without a leader is shut down.
+are shut down without notifying the leaders upfront. In this case, it takes Spanner
+about 10s to recover. If the leaders are notified to abdicate before their zone
+is shut down, the availability is almost unaffected. The same is true if a zone
+without a leader is shut down.
 
-Google also provided results for some experiments concerning TrueTime. All
-assumptions Google made when designing Spanner would break if the local clock
-drift would exceed $200\mu$s/sec. But Google found that clock issues are extremely
-infrequent relative to more serious hardware problems, e.g. bad CPUs. Moreover,
-Google provided empirical results on the values of $\epsilon$. Typically,
-$\epsilon$ is never larger than 10ms. It decreases as the network infrastructure
-is improved and increases if time masters go offline.
+Google also provided empirical results about TrueTime. All assumptions Google
+made while designing Spanner would be broken if the local clock drift of a machine
+exceeds $200\mu$s/sec. But Google found that clock issues are extremely infrequent
+relative to more serious hardware problems, e.g. bad CPUs. Moreover, Google
+provided empirical results on the values of $\epsilon$. Typically, $\epsilon$ is
+never larger than 10ms. It decreases as the network infrastructure is improved
+and increases if time masters go offline.
 

### 2016-02-16 paper/paper.tex
Iterate Future Work

@@
 
-\section{Future Work \tbr}
+\section{Future Work}
 \label{sec:future-work}
 
-Google plans to implement optimistic parallel reads but it turns out that
-implementing this is non-trivial. They also want to support direct configuration
-changes of the Paxos state machines and to reduce $\epsilon$ to be $\leq 1ms$ to
-improve the performance if data is sharded between nearby datacenters.
+Google plans to implement optimistic parallel reads but it turned out that
+implementing it is non-trivial. They also want to support direct configuration
+changes of the Paxos state machines and to reduce $\epsilon$ to be $\leq 1ms$ in
+order to improve the performance if data is sharded between nearby datacenters.
 
-Moreover, the node-local performance can be improved for complex queries. By now,
-node-local data structures are designed for key-value accesses and these could be
-improved by algorithms and data structure from database literature.
+Moreover, the node-local performance could be improved for complex queries. By
+now, node-local data structures are designed for key-value accesses and these data
+structures could be improved by algorithms and data structure from database
+literature.
 
 To improve the performance even further, Google plans to move processes between
-datacenters if the data they operate on is moved. This would have a high impact
-because the performance of the processes is close to optimal if the data is as
-local as possible. However, moving processes between datacenters raises even more
-complex problems.
+datacenters if the data on which they operate is moved. This would have high
+impact because the performance of the processes is close to optimal if the data
+is as local as possible. However, moving processes between datacenters raises
+even more complex problems.
 

### 2016-02-16 paper/paper.tex
Iterate Discussion

@@
 
-\section{Discussion \tbr}
+\section{Discussion}
 \label{sec:discussion}
@@
 to establish a notion of global wall clock time (that's how they call it in their
-presentation) which enabled them to guarantee serialization order for
-globally-distributed transactions. To the best of the author's knowledge, Spanner
+presentation). Global wall clock time enables them to guarantee serialization order
+for globally-distributed transactions. To the best of the author's knowledge, Spanner
 was the first system to provide much stronger consistency and time semantics while
@@
 developer's point of view, this is interesting because Spanner provides
-the more intuitive schematized view of tables. In contrast to NOSQL database
-systems, Spanner could be used as replacement for other relational database
+the more intuitive schematized view of tables. In contrast to NoSQL database
+systems, Spanner can be used as replacement for other relational database
 systems, e.g. MySQL, directly. Spanner is also highly flexible to schema changes
-because of the atomic schema change transactions.
+because it supports atomic schema change transactions.
 
@@
 example all the GPS and atomic clocks. Unless this hardware infrastructure is
-not available it is also hard to reimplement the TrueTime API. Nonetheless,
+available it is hard to reimplement the TrueTime API. Nonetheless,
 companies which need a globally-distributed database with all the guarantees
 Spanner provides are likely to be almost as big as Google. Hence, such companies
-may also be able to build a Spanner-like system. If every datacenter would provide
-GPS and Armageddon time masters independently, the reimplementation of such a
-system would be much easier.
+would also have the resources to build a Spanner-like system. An Open-Source
+reimplementation would only be possible if every datacenter would provide GPS
+and Armageddon time masters on their own.
 
-Another part that may influence the performance of Spanner is the distributed
-file system, i.e. Colossus, the successor of the Google File System (GFS). I did
-not find reliable information about the implementation or the performance of
-Colossus. It is unknown how much impact this distributed file system has on the
+A component that may influence the performance of Spanner is the distributed
+file system, i.e. Colossus, the successor of the Google File System. The internet 
+does not provide reliable information about the implementation or the performance
+of Colossus. It is unknown how much impact this distributed file system has on the
 performance of Spanner. Thus, it is hard to say if a reimplementation of Spanner
-would be able to compete with Spanner's performance.
+would be able to compete with Spanner's performance if a different distributed
+file system is used.
 
-Of course, such a system may also be built upon e.g. the Network Time Protocol
-(NTP). But this would result in poor performance compared to Spanner because the
-error interval would be much larger compared to $2\cdot\epsilon$. And besides
-providing distributed general-purpose transactions, the performance of Spanner
-makes the whole system that impressive. A system using e.g. NTP would never be
-able to achieve this performance while guaranteeing serialization order. Spanner
-can synchronize multiple systems without any communication between these systems.
-NTP-based systems would have to communicate with each other. For systems far apart
-such a system would never be able to compete with Spanner.
+Of course, a distributed database system may also be built upon other concepts,
+e.g. the Network Time Protocol (NTP). But this would result in poor performance
+compared to Spanner because the error interval would be much larger then the one
+used by Spanner, i.e. $2\cdot\epsilon$. And besides providing distributed
+general-purpose transactions, it is the performance of Spanner that impressives
+the most. A system using NTP would never be able to achieve this performance
+while guaranteeing serialization order of transactions. Spanner can synchronize
+multiple systems without any communication between these systems. NTP-based
+systems would have to communicate with each other. For systems far apart such a
+NTP-based solution would never be able to compete with Spanner.
 
-Spanner is robust to GPS blackouts because of the usage ofatomic clocks, even
+Spanner is robust to GPS blackouts because of the usage of atomic clocks, even
 though the performance of Spanner may suffer due to high values of $\epsilon$.
-Would be interesting to see how such a blackout affects Spanner's performance.
+Would be interesting to see how a GPS blackout would affect Spanner's performance.
 
-It would also be interesting how failures of the two singletons per universe,
-i.e. universemaster and placement driver, affect Spanner. Since the universemaster
-is just a debugging console, most probably a failure in this component will not
-affect Spanner at all. This would probably be a totally different story if the
-placement driver does not operate correctly. First of all, there would be no
-load balancing between zones. Hence, zones could be run out of space if the
-placement driver is down long enough. Furthermore, updated replication
-constraints, would not be propagated since the placement driver is responsible
-to move the data if the replication constraints are modified.
+It would also be interesting to know how failures of the two singletons per
+universe, i.e. universemaster and placement driver, would affect Spanner. Since
+the universemaster is just a debugging console, most probably a failure in this
+component will not affect Spanner at all. This would probably be a totally
+different story if the placement driver does not operate correctly. First of all,
+there would be no load balancing between zones. Hence, zones could run out of
+space if the placement driver is down long enough. Furthermore, updated
+replication constraints, would not be propagated since the placement driver is
+responsible to move the data if the replication constraints are modified.
+Therefore I do not understand why there is no redundancy in placement driver.
 
@@
 lots of replicas which are spread across the globe and writes large amounts of data
-regularly, Spanner may not provide the performance mentioned in the original
-paper. However, Spanner may still outperform both, Megastore and Bigtable while
-providing much stronger consistency guarantees.
+regularly, Spanner may not provide the performance presented in the original
+publication. However, Spanner may still outperform all other distributed database
+systems while providing much stronger consistency guarantees.
 

### 2016-02-25 paper/paper.tex
Remove tbr
Iterate section 1 and 2

@@
 
-Google already has two other highly scalable distributed databases, i.e.
+Google already has two other highly scalable distributed databases, i.e.,
 Bigtable~\cite{Chang:2008} and Megastore~\cite{Baker:2011}. Hence, there must be
@@
 
-\subsection{Spanserver Software Stack \tbr}
+\subsection{Spanserver Software Stack}
 \label{subsec:spanserver-software-stack}
@@
 On top of each tablet, a Paxos state machine is implemented in order to support
-consistent replication. One Paxos state machine is elect as leader. Leader leases
-are assigned in a long-lived manner (default lease length is 10 seconds). A leader
-implements three additional layers. A \emph{lock table} which stores the state
-for two-phase locking and therefore implements the concurrency control. On top
-of the lock table, a \emph{transaction manager} is implemented to coordinate
+consistent replication. One Paxos state machine is elected as leader. Leader
+leases are assigned in a long-lived manner (default lease length is 10 seconds).
+A leader implements three additional layers. A \emph{lock table} which stores the
+state for two-phase locking and therefore implements the concurrency control. On
+top of the lock table, a \emph{transaction manager} is implemented to coordinate
 distributed transactions. In turn, this transaction manager is used to implement
@@
 involves multiple Paxos groups (set of replicas). In this case, one of the
-participant groups is chosen as coordinator and the leader of this group is
+participant groups is chosen as a coordinator and the leader of this group is
 referred to as \emph{coordinator leader}.
@@
 
+\begin{figure}[ht]
+  \centering
+  \begin{subfigure}{0.45\textwidth}
+    \includegraphics[width = \textwidth]{figs/directories}
+    \caption{Directories (unit of data movement).~\cite{Corbett:2012}}
+    \label{subfig:directories}
+  \end{subfigure}
+  \qquad
+  \begin{subfigure}{0.45\textwidth}
+    \includegraphics[width = \textwidth]{figs/interleaving-example.png}
+    \caption{Two interleaved tables.~\cite{Corbett:2012}}
+    \label{subfig:interleaved-tables}
+  \end{subfigure}
+  \caption{Directories and interleaved tables.}
+  \label{fig:directories-and-tables}
+\end{figure}
+
 All the data within a directory has the same replication configuration. As will
@@
 mapping of Spanner and the one used in Bigtable. A Spanner tablet may contain
-multiple contiguous ranges of keys (= directories) as opposed to a Bigtable tablet
-which may only contain one contiguous range of keys.
-Figure~\ref{fig:comparison-tablets} shows an example comparison of the two tablet
-types.
+multiple contiguous ranges of keys (= directories) with gaps between contiguous
+ranges as opposed to a Bigtable tablet which may only contain one contiguous key
+range. Figure~\ref{fig:comparison-tablets} shows an example comparison of the two
+tablet types.
 
@@
 application/user can specify geographic-replication properties. An application
-can control replication in two dimensions, namely the number and types of
+can control replication along two dimensions, namely the number and types of
 replicas, and the geographic placement of replicas.
-
-\begin{figure}[ht]
-  \centering
-  \begin{subfigure}{0.45\textwidth}
-    \includegraphics[width = \textwidth]{figs/directories}
-    \caption{Directories (unit of data movement).~\cite{Corbett:2012}}
-    \label{subfig:directories}
-  \end{subfigure}
-  \qquad
-  \begin{subfigure}{0.45\textwidth}
-    \includegraphics[width = \textwidth]{figs/interleaving-example.png}
-    \caption{Two interleaved tables.~\cite{Corbett:2012}}
-    \label{subfig:interleaved-tables}
-  \end{subfigure}
-  \caption{Directories and interleaved tables.}
-  \label{fig:directories-and-tables}
-\end{figure}
 
@@
 
-There exists another layer on top of the directory-bucketed key-value mappings.
-This layer is probably the most intuitive one for people who are used to common
-database systems. The data model allows applications to create databases in a
-universe each of which may contain any number of schematized
-\emph{semi-relational} tables. A table has rows, columns and, as a third
-dimension, versioned values. Spanner's tables are semi-relational because rows
-must have names, i.e. every table must have an ordered set of one or more
-primary-key columns. This is useful for applications because the locality of
-data can be controlled by choosing keys accurately.
+There exists another layer on top of the directory-bucketed key-value mappings,
+i.e., schematized \emph{semi-relational} tables. This layer is probably the most
+intuitive one for people who are used to common database systems. The data model
+allows applications to create databases in a universe each of which may contain
+an arbitrary number of schematized semi-relational tables. A table has rows,
+columns, and (as a third dimension) versioned values. Spanner's tables are
+semi-relational because rows must have names, i.e., every table must have an
+ordered set of one or more primary-key columns. This is useful for applications
+because the locality of data can be controlled by choosing keys accurately.
 
@@
 rows of \texttt{Albums} connected to user \texttt{1} are located in the same
-directory (\texttt{3365}). This directory has a common prefix, i.e. \texttt{1},
-from the \texttt{Users} table. The same principle can be seen for key \texttt{2}
+directory (\texttt{3365}). This directory has a common prefix, \texttt{1},
+from the \texttt{Users} table. The same principle applies to key \texttt{2}
 of the \texttt{Users} table. This interleaving of tables exploits locality
@@
 
-This section describes the key enabler of Spanner, i.e. the novel \emph{TrueTime}
+This section describes the key enabler of Spanner, i.e., the novel \emph{TrueTime}
 API. Essentially, this API enables Google to assign timestamps to distributed
@@
 TrueTime represents time as a $TTinterval$. A $TTinterval$ is an interval with a
-bounded time uncertainty and has two endpoints, i.e. $earliest$ and $latest$,
+bounded time uncertainty and has two endpoints, i.e., $earliest$ and $latest$,
 which are of type $TTstamp$ (which is a timestamp in the context of Spanner).
@@
 which is guaranteed to contain the absolute time of the invocation of $TT.now()$,
-i.e. $TT.now().earliest \leq t_{abs}(e) \leq TT.now().latest$, where $e$ denotes
+i.e., $TT.now().earliest \leq t_{abs}(e) \leq TT.now().latest$, where $e$ denotes
 the event of invoking $TT.now()$. More precisely, the absolute time of invocation
@@
 The TrueTime API allows Spanner to guarantee some correctness properties of the
-implemented concurrency control. These enable some of the main features, i.e.
+implemented concurrency control. These enable some of the main features, i.e.,
 externally-consistent transactions, lock-free read-only transactions and so-called
@@
 Read-write transactions use two-phase locking and Spanner may only assign
-timestamps in between the two phases. Spanner enforces another invariant, i.e. a
-\emph{monotonicity invariant}, which makes use of the disjointness invariant.
+timestamps in between the two phases. Spanner enforces another invariant, i.e.,
+a \emph{monotonicity invariant}, which makes use of the disjointness invariant.
 This monotonicity invariant guarantees that within each Paxos group, Spanner
@@
 Figure~\ref{fig:read-write-transaction} illustrates the timeline of a distributed
-read-write transaction. There are three participants groups involved, i.e. $P_2$,
+read-write transaction. There are three participants groups involved, i.e., $P_2$,
 $P_3$ and $P_{CL}$ (which is the coordinator leader). In the first step, all three
@@
 The safe time is determined by the safe time of a Paxos state machine, denoted
-as $t_{safe}^{Paxos}$, and the safe time of the transaction manager, i.e.
+as $t_{safe}^{Paxos}$, and the safe time of the transaction manager, i.e.,
 $t_{safe}^{TM}$. The overall safe time of a replica then is the minimum of
@@
 
-This section summarizes the core results of the evaluation of Spanner, i.e. the
+This section summarizes the core results of the evaluation of Spanner, i.e., the
 latency and throughput performance as well as some results concerning availability
@@
 A component that may influence the performance of Spanner is the distributed
-file system, i.e. Colossus, the successor of the Google File System. The internet 
+file system, i.e., Colossus, the successor of the Google File System. The internet 
 does not provide reliable information about the implementation or the performance
@@
 compared to Spanner because the error interval would be much larger then the one
-used by Spanner, i.e. $2\cdot\epsilon$. And besides providing distributed
+used by Spanner, i.e., $2\cdot\epsilon$. And besides providing distributed
 general-purpose transactions, it is the performance of Spanner that impressives
@@
 It would also be interesting to know how failures of the two singletons per
-universe, i.e. universemaster and placement driver, would affect Spanner. Since
+universe, i.e., universemaster and placement driver, would affect Spanner. Since
 the universemaster is just a debugging console, most probably a failure in this

### 2016-02-25 paper/paper.tex
Iterate section 3

@@
 TrueTime represents time as a $TTinterval$. A $TTinterval$ is an interval with a
-bounded time uncertainty and has two endpoints, i.e., $earliest$ and $latest$,
-which are of type $TTstamp$ (which is a timestamp in the context of Spanner).
+bounded time uncertainty and has two endpoints, $earliest$ and $latest$, which
+are of type $TTstamp$ (a timestamp in the context of Spanner).
 Table~\ref{tbl:truetime-api} summarizes the three methods provided by the TrueTime
-API. However, the most important method is $TT.now()$ which returns a $TTinterval$
-which is guaranteed to contain the absolute time of the invocation of $TT.now()$,
-i.e., $TT.now().earliest \leq t_{abs}(e) \leq TT.now().latest$, where $e$ denotes
-the event of invoking $TT.now()$. More precisely, the absolute time of invocation
-of $TT.now()$ is exactly in the middle this $TTinterval$. The interval itself
-spans over $2\cdot\epsilon$, where $\epsilon$ denotes the instantaneous error
-bound. $\epsilon$ is derived from applied worst-case local clock drift (assumed
-to be $200\mu$s/sec). This instantaneous error bound is adjusted over time for
-every single machine.
+API. However, the most important method is $TT.now()$. i$TT.now()$ which returns
+a $TTinterval$ which is guaranteed to contain the absolute time of the invocation
+of $TT.now()$, i.e., $TT.now().earliest \leq t_{abs}(e) \leq TT.now().latest$,
+where $e$ denotes the event of invoking $TT.now()$. More precisely, the absolute
+time of invocation of $TT.now()$ is exactly in the middle this $TTinterval$. The
+interval itself spans over $2\cdot\epsilon$, where $\epsilon$ denotes the
+instantaneous error bound. $\epsilon$ is derived from applied worst-case local
+clock drift (assumed to be $200\mu$s/sec). This instantaneous error bound is
+adjusted over time for every single machine.
 
@@
 types of time references because the failure modes of them are uncorrelated. A
-datacenter typically consists of multiple \emph{time master} machines. Most of
+datacenter typically maintains multiple \emph{time master} machines. Most of
 these are GPS receivers with dedicated antennas. All remaining time master

### 2016-02-25 paper/paper.tex
Iterate section 4

@@
 The TrueTime API allows Spanner to guarantee some correctness properties of the
-implemented concurrency control. These enable some of the main features, i.e.,
-externally-consistent transactions, lock-free read-only transactions and so-called
-\emph{snapshot reads}. Snapshot reads are non-blocking reads in the past at a
-specific timestamp $t$. In this case, Spanner guarantees that snapshot reads will
+implemented concurrency control. These enable some of the main features:
+externally-consistent transactions, lock-free read-only transactions, and
+so-called \emph{snapshot reads}. Snapshot reads are non-blocking reads in the
+past at a specific timestamp $t$. Spanner guarantees that snapshot reads will
 see the effects of all transactions committed as of $t$.
 
-Spanner provides different operations, as can be seen in
-Table~\ref{tbl:reads-writes}. Read-only transactions have to be made explicit.
-This requires the application/user to predeclare that the read-only transaction
-does not include write operations. Moreover, read-only transaction have the
-benefits of snapshot isolation because they are \emph{lock-free}. Standalone
-writes are simply read-write transaction without any read operations. Non-snapshot
-standalone reads are implemented as read-only transactions. Read-only transactions
-and snapshot reads proceed on any \emph{sufficiently up-to-date} replica. For
-snapshot reads, the application/user has two options. On the one hand, a precise
-timestamp can be specified at which the read should be executed. On the other
-hand, an upper bound can be specified and Spanner choses a timestamp that
-satisfies this bound.
+Spanner provides different operations, as shown in Table~\ref{tbl:reads-writes}.
+Read-only transactions have to be made explicit. This requires the
+application/user to predeclare that the read-only transaction does not include
+write operations. Moreover, read-only transaction have the benefits of snapshot
+isolation because they are \emph{lock-free}. Standalone writes are simply
+read-write transaction without any read operations. Non-snapshot standalone reads
+are implemented as read-only transactions. In contrast to snapshot reads, which
+may be executed on any timestamp in the past, read-only transactions execute on
+the last committed timestamp. Both, snapshot reads and read-only transactions,
+proceed on any \emph{sufficiently up-to-date} replica. For snapshot reads,
+the application/user has two options. On the one hand, a precise timestamp can
+be specified at which the read should be executed. On the other hand, an upper
+bound can be specified, and Spanner choses a timestamp that satisfies this bound.
 
-\begin{table}
+\begin{table}[ht]
   \centering
@@
 
-Read-write transactions use two-phase locking and Spanner may only assign
-timestamps in between the two phases. Spanner enforces another invariant, i.e.,
+Read-write transactions use two-phase locking, and Spanner may only assign
+timestamps in between the two phases. Spanner enforces another invariant, the
 a \emph{monotonicity invariant}, which makes use of the disjointness invariant.
@@
 \emph{external consistency invariant}. In essence, this invariant guarantees that
-whenever the commit of a transaction $T_1$, denoted as $e_1^{commit}$, occurs
-before the start of a transaction $T_2$, denoted as $e_2^{start}$, then the commit
-timestamp of $T_1$, denoted as $s_1$, is guaranteed to be smaller than the
-timestamp of $T_2$, denoted as $s_2$. More formally, the following must hold to
-guarantee external consistency:
+whenever the commit of a transaction $T_1$, denoted as $e_1^{commit}$, in real
+time occurs before the start of a transaction $T_2$, denoted as $e_2^{start}$,
+then the commit timestamp of $T_1$, denoted as $s_1$, is guaranteed to be smaller
+than the commit timestamp of $T_2$, denoted as $s_2$. More formally, the following
+must hold to guarantee external consistency:
 \begin{center}
@@
 \begin{description}
-  \item[Start] Coordinator leader for a write $T_i$ assigns a commit timestamp
-    $s_i \geq TT.now().latest$, where $TT.now()$ is executed right after the
-    arrival event of the commit request for $T_i$.
+  \item[Start] Coordinator leader for a write transaction $T_i$ assigns a commit
+    timestamp $s_i \geq TT.now().latest$, where $TT.now()$ is executed right after
+    the arrival event of the commit request for $T_i$.
   \item[Commit Wait] Coordinator leader ensures that clients cannot see data

### 2016-02-25 paper/paper.tex
Iterate section 5

@@
 Figure~\ref{fig:read-write-transaction} illustrates the timeline of a distributed
-read-write transaction. There are three participants groups involved, i.e., $P_2$,
-$P_3$ and $P_{CL}$ (which is the coordinator leader). In the first step, all three
+read-write transaction. There are three participants groups involved: $P_2$, $P_3$
+and $P_{CL}$ (which is the coordinator leader). In the first step, all three
 participants acquire their locks and execute $TT.now()$ right afterwards to obtain
@@
 replicas because the number of spanservers increases. In the experiments, the
-number of replicas and the number of spanservers were equal. The throughput of write
-transactions also benefits from this fact. However, the number of write operations
-to perform increases linearly with the number of replicas. This explains why the
-throughput of write transactions does not increase with the number of replicas.
-Snapshot reads benefit the most from the Spanner architecture. The throughput of
-snapshot reads increases almost linearly with the number of replicas.
+number of replicas and the number of spanservers were equal. The throughput of
+write transactions also benefits from this fact. However, the number of write
+operations to be performed increases linearly with the number of replicas. This
+explains why the throughput of write transactions does not increase with the
+number of replicas. Snapshot reads benefit the most from the Spanner architecture.
+The throughput of snapshot reads increases almost linearly with the number of
+replicas.
 

### 2016-02-25 paper/paper.tex
Iterate section 7

@@
 presentation). Global wall clock time enables them to guarantee serialization order
-for globally-distributed transactions. To the best of the author's knowledge, Spanner
-was the first system to provide much stronger consistency and time semantics while
-data is sharded and replicated at global scale. Google also proved all
-'\emph{Relational databases are dead}' people wrong even though systems like
+for globally-distributed transactions. To the best of the author's knowledge,
+Spanner was the first system to provide much stronger consistency and time
+semantics while data is sharded and replicated at global scale. Google also proved
+all '\emph{Relational databases are dead}' people wrong even though systems like
 Bigtable helped creating the NoSQL trend. Reading about a system like Spanner is
@@
 small- to mid-sized companies because there is a lot of hardware involved, for
-example all the GPS and atomic clocks. Unless this hardware infrastructure is
+example, all the GPS and atomic clocks. Unless this hardware infrastructure is
 available it is hard to reimplement the TrueTime API. Nonetheless,
@@
 there would be no load balancing between zones. Hence, zones could run out of
-space if the placement driver is down long enough. Furthermore, updated
-replication constraints, would not be propagated since the placement driver is
-responsible to move the data if the replication constraints are modified.
-Therefore I do not understand why there is no redundancy in placement driver.
+space if the placement driver is down for an extended period of time. Furthermore,
+updated replication constraints, would not be propagated since the placement
+driver is responsible for moving the data if the replication constraints are
+modified. Therefore, I do not understand why there is no redundancy in placement
+drivers.
 

### 2016-02-25 paper/paper.tex
Iterate section 2
Iterate section 3

@@
 an arbitrary number of schematized semi-relational tables. A table has rows,
-columns, and (as a third dimension) versioned values. Spanner's tables are
-semi-relational because rows must have names, i.e., every table must have an
-ordered set of one or more primary-key columns. This is useful for applications
-because the locality of data can be controlled by choosing keys accurately.
+columns, and versioned values. Spanner's tables are semi-relational because rows
+must have names, i.e., every table must have an ordered set of one or more
+primary-key columns. This is useful for applications because the locality of data
+can be controlled by choosing keys accurately.
 
@@
 Table~\ref{tbl:truetime-api} summarizes the three methods provided by the TrueTime
-API. However, the most important method is $TT.now()$. i$TT.now()$ which returns
-a $TTinterval$ which is guaranteed to contain the absolute time of the invocation
+API. However, the most important method is $TT.now()$. $TT.now()$ returns a
+$TTinterval$ which is guaranteed to contain the absolute time of the invocation
 of $TT.now()$, i.e., $TT.now().earliest \leq t_{abs}(e) \leq TT.now().latest$,

### 2016-02-25 paper/paper.tex
Iterate section 4

@@
 may be executed on any timestamp in the past, read-only transactions execute on
-the last committed timestamp. Both, snapshot reads and read-only transactions,
-proceed on any \emph{sufficiently up-to-date} replica. For snapshot reads,
-the application/user has two options. On the one hand, a precise timestamp can
-be specified at which the read should be executed. On the other hand, an upper
+the last committed write timestamp. Both, snapshot reads and read-only
+transactions, proceed on any \emph{sufficiently up-to-date} replica. For snapshot
+reads, the application/user has two options. On the one hand, a precise timestamp
+can be specified at which the read should be executed. On the other hand, an upper
 bound can be specified, and Spanner choses a timestamp that satisfies this bound.
@@
 timestamps in between the two phases. Spanner enforces another invariant, the
-a \emph{monotonicity invariant}, which makes use of the disjointness invariant.
+\emph{monotonicity invariant}, which makes use of the disjointness invariant.
 This monotonicity invariant guarantees that within each Paxos group, Spanner
